---
title: "Mini-Project 01 - Assessing the Impact of SFFA on Campus Diversity One-Year Later"
author: "dhwanikan"
editor:
    mode: source
format:
    html:
        code-fold: true
---

## Introduction

Nobody knew exactly what would happen. That was the honest answer in 
June 2023, when the Supreme Court struck down race-conscious admissions 
in *Students for Fair Admissions v. Harvard*. Advocates on both sides 
made confident predictions --- catastrophic drops in Black and Hispanic 
enrollment on one hand, barely a ripple on the other. The truth, as 
usual, was going to require actual data.

That data is now here. Fall 2024 marks the first full admissions cycle 
conducted entirely under post-SFFA rules, and IPEDS --- the federal 
clearinghouse for college enrollment statistics --- has released the numbers.

This project works through those numbers systematically. Using 15 years 
of enrollment records (2010--2024), I examine how racial composition has 
shifted at institutions across the country, paying close attention to 
CUNY schools and California's public universities. The California schools 
are a natural control group: Proposition 209 banned affirmative action 
there in 1996, so their demographic trends tell us what campuses look 
like when admissions have been race-blind for decades. Comparing other 
institutions against that benchmark is the cleanest way to separate 
genuine SFFA effects from background noise.

## Data Acquisition and Preparation

### Task 1: Acquire IPEDS Data

The following function downloads fall enrollment (EF) and institution 
description (HD) files from IPEDS for 2010--2024. Files are cached locally 
in `data/mp01/` to avoid repeated downloads.

```{r}
#| label: acquire-data
#| message: false
#| warning: false
#| cache: true

library(tidyverse)
library(glue)

acquire_ipeds_data <- function(start_year=2010, end_year=2024){
    data_dir <- file.path("data", "mp01")
    if(!dir.exists(data_dir)){
        dir.create(data_dir, showWarnings=FALSE, recursive=TRUE)
    }
    YEARS <- seq(start_year, end_year)

    EFA_ALL <- map(YEARS, function(yy){
        if(yy <= 2022){
            ef_url <- glue("https://nces.ed.gov/ipeds/datacenter/data/EF{yy}A.zip")
        } else {
            ef_url <- glue("https://nces.ed.gov/ipeds/data-generator?year={yy}&tableName=EF{yy}A&HasRV=0&type=csv")
        }
        ef_file <- file.path(data_dir, glue("ef{yy}a.csv.zip"))
        if(!file.exists(ef_file)){
            message(glue("Downloading Enrollment Data for {yy}"))
            download.file(ef_url, destfile=ef_file, quiet=TRUE, mode="wb")
        }
        read_csv(ef_file, show_col_types=FALSE) |>
            mutate(year = yy,
                   enrollment_m_aian = EFAIANM, enrollment_f_aian = EFAIANW,
                   enrollment_m_asia = EFASIAM, enrollment_f_asia = EFASIAW,
                   enrollment_m_bkaa = EFBKAAM, enrollment_f_bkaa = EFBKAAW,
                   enrollment_m_hisp = EFHISPM, enrollment_f_hisp = EFHISPW,
                   enrollment_m_nhpi = EFNHPIM, enrollment_f_nhpi = EFNHPIW,
                   enrollment_m_whit = EFWHITM, enrollment_f_whit = EFWHITW,
                   enrollment_m_2mor = EF2MORM, enrollment_f_2mor = EF2MORW,
                   enrollment_m_unkn = EFUNKNM, enrollment_f_unkn = EFUNKNW,
                   enrollment_m_nral = EFNRALM, enrollment_f_nral = EFNRALW) |>
            filter((EFALEVEL %in% c(2, 12)) | (LINE %in% c(1, 15))) |>
            mutate(level = case_when(
                EFALEVEL == 2  ~ "all undergrad",
                EFALEVEL == 12 ~ "all graduate",
                LINE %in% c(1, 15) ~ "first year undergrad"
            )) |>
            select(institution_id = UNITID, year, level,
                   starts_with("enrollment_")) |>
            group_by(institution_id, year, level) |>
            summarize(across(starts_with("enrollment_"), sum), .groups="drop")
    }) |> bind_rows()

    DESC_ALL <- map(YEARS, function(yy){
        if(yy <= 2022){
            hd_url <- glue("https://nces.ed.gov/ipeds/datacenter/data/HD{yy}.zip")
        } else {
            hd_url <- glue("https://nces.ed.gov/ipeds/data-generator?year={yy}&tableName=HD{yy}&HasRV=0&type=csv")
        }
        hd_file <- file.path(data_dir, glue("hd{yy}.csv.zip"))
        if(!file.exists(hd_file)){
            message(glue("Downloading Institution Descriptions for {yy}"))
            download.file(hd_url, destfile=hd_file, quiet=TRUE, mode="wb")
        }
        suppressWarnings(
            read_csv(hd_file, show_col_types=FALSE,
                     locale=locale(encoding=if_else(yy==2024,"utf-8","windows-1252"))) |>
            mutate(year=yy) |>
            select(institution_id=UNITID, institution_name=INSTNM,
                   state=STABBR, year)
        )
    }) |> bind_rows()

    inner_join(EFA_ALL, DESC_ALL, join_by(institution_id, year))
}

IPEDS <- acquire_ipeds_data()
```

### Tasks 2 & 3: Add Helper Columns

```{r}
#| label: clean-data

# Add total enrollment columns and institutional flags
IPEDS <- IPEDS |>
    mutate(
        # Row totals --- used throughout the analysis
        enrollment_total   = rowSums(across(starts_with("enrollment_"))),
        enrollment_f_total = rowSums(across(starts_with("enrollment_f_"))),
        enrollment_m_total = rowSums(across(starts_with("enrollment_m_"))),

        # Task 2: CUNY flag --- TRUE if "CUNY" appears in the institution name
        is_cuny = str_detect(institution_name, "CUNY"),

        # Task 3: California public flag --- UC system OR Cal State system
        # Must use Boolean logic (|), not a single regex, per assignment instructions
        is_calpublic = str_detect(institution_name, "University of California") |
                       str_detect(institution_name, "California State")
    )
```

Verify that the California public flag is working correctly by checking
which institutions it identifies:

```{r}
#| label: verify-calpublic

IPEDS |>
    filter(is_calpublic) |>
    distinct(institution_name) |>
    arrange(institution_name)
```

## Exploratory Analysis

### Task 4: Initial EDA Pass

Before diving into specific questions, two quick sanity checks confirm the
data loaded correctly and has the expected structure.

**Check 1 --- Column types:**

```{r}
#| label: glimpse-data

glimpse(IPEDS)
```

All enrollment columns are numeric, `is_cuny` and `is_calpublic` are logical
(TRUE/FALSE), `year` is an integer, and `institution_name` and `state` are
character columns. No unexpected type coercions occurred.

**Check 2 --- Summary statistics:**

```{r}
#| label: skim-data

library(skimr)
skim(IPEDS)
```

The dataset spans 2010--2024 and covers a large number of institutions.
Enrollment values are non-negative throughout. The two Boolean flag columns
have no missing values. Some rows show zero enrollment in certain demographic
categories, which is expected for smaller institutions with limited enrollment
in a given group.

---

### Task 5: Inline Value Questions

```{r}
#| label: task5-calculations
#| message: false

library(scales)

# Q1: Distinct institution count
n_institutions <- n_distinct(IPEDS$institution_id)

# Q2: Baruch graduate students in 2024
baruch_grad_2024 <- IPEDS |>
    filter(str_detect(institution_name, "Baruch"),
           level == "all graduate",
           year == 2024) |>
    summarize(total = sum(enrollment_total)) |>
    pull(total)

# Q3: Total Baruch students in 2024
# Use only "all undergrad" + "all graduate" to avoid double-counting
# first-year students, who are already a subset of "all undergrad"
baruch_total_2024 <- IPEDS |>
    filter(str_detect(institution_name, "Baruch"),
           level %in% c("all undergrad", "all graduate"),
           year == 2024) |>
    summarize(total = sum(enrollment_total)) |>
    pull(total)

# Q4: Institution with most female undergraduates in 2019
most_female_2019 <- IPEDS |>
    filter(year == 2019, level == "all undergrad") |>
    group_by(institution_name) |>
    summarize(total_female = sum(enrollment_f_total), .groups = "drop") |>
    slice_max(total_female, n = 1)

most_female_inst  <- most_female_2019$institution_name
most_female_count <- most_female_2019$total_female

# Q5: Highest NHPI fraction among 1,000+ first-year students in 2024
nhpi_top <- IPEDS |>
    filter(year == 2024, level == "first year undergrad") |>
    mutate(nhpi_total = enrollment_m_nhpi + enrollment_f_nhpi) |>
    filter(enrollment_total > 1000) |>
    mutate(nhpi_frac = nhpi_total / enrollment_total) |>
    slice_max(nhpi_frac, n = 1)

nhpi_inst <- nhpi_top$institution_name
nhpi_pct  <- nhpi_top$nhpi_frac
```

The IPEDS dataset contains **`r comma(n_institutions)`** distinct institutions.
At Baruch College in 2024, **`r comma(baruch_grad_2024)`** graduate students
were enrolled, and **`r comma(baruch_total_2024)`** students in total. The
institution with the most enrolled female undergraduates in 2019 was
**`r most_female_inst`**, with **`r comma(most_female_count)`** female students.
Among institutions enrolling over 1,000 first-year undergraduates in 2024,
**`r nhpi_inst`** had the highest proportion of Native Hawaiian or Pacific
Islander students at **`r percent(nhpi_pct, accuracy=0.1)`**.

```{r}
#| label: covid-dip-check
#| message: false

# Check whether the 2020-2021 COVID dip affects our baseline comparison.
# This is an important methodological check before attributing any 2024
# shifts to SFFA --- COVID caused its own demographic disruptions.
baruch_covid <- IPEDS |>
    filter(str_detect(institution_name, "Baruch"),
           level == "first year undergrad",
           year %in% c(2019, 2020, 2021, 2022)) |>
    mutate(
        minority = enrollment_m_bkaa + enrollment_f_bkaa +
                   enrollment_m_hisp + enrollment_f_hisp +
                   enrollment_m_aian + enrollment_f_aian +
                   enrollment_m_nhpi + enrollment_f_nhpi +
                   enrollment_m_2mor + enrollment_f_2mor
    ) |>
    group_by(year) |>
    summarize(
        total     = sum(enrollment_total),
        diversity = sum(minority) / sum(enrollment_total),
        .groups   = "drop"
    )

pre_covid_div <- baruch_covid |> filter(year == 2019) |> pull(diversity)
covid_div     <- baruch_covid |> filter(year == 2020) |> pull(diversity)
recovery_div  <- baruch_covid |> filter(year == 2022) |> pull(diversity)
covid_drop    <- covid_div - pre_covid_div
```

One important methodological note before proceeding: before attributing any
2024 demographic shift to SFFA, we should account for the COVID-19 disruption.
Baruch's first-year diversity `r if_else(covid_drop >= 0, "rose", "fell")` by
`r percent(abs(covid_drop), accuracy=0.1)` between 2019 and 2020 as the
pandemic reshaped who applied and enrolled. By 2022, diversity had
`r if_else(recovery_div > pre_covid_div, "recovered and exceeded", "nearly returned to")`
pre-pandemic levels at `r percent(recovery_div, accuracy=0.1)`. This COVID
baseline effect means that simple 2023-to-2024 comparisons may conflate
pandemic recovery with genuine SFFA effects --- a subtlety the California
comparison helps control for, since California schools experienced the same
COVID disruption without any change in admissions policy.

These figures begin to sketch the contours of American higher education --- a
landscape dominated by a handful of large states, with CUNY serving as a
microcosm of New York City's own demographic complexity. The year-over-year
fluctuations at Baruch hint at broader enrollment pressures that the following
tables explore in more detail.

---

### Task 6: Exploratory Tables

```{r}
#| label: task6-table1
#| message: false

library(gt)
library(gtExtras)

# Table 1: Top 5 states by total graduate enrollment
IPEDS |>
    filter(level == "all graduate") |>
    group_by(state) |>
    summarize(grad_students = sum(enrollment_total), .groups = "drop") |>
    slice_max(grad_students, n = 5) |>
    arrange(desc(grad_students)) |>
    gt() |>
    tab_header(
        title    = "Top 5 States by Graduate Enrollment",
        subtitle = "Cumulative enrollment across all institutions, 2010-2024"
    ) |>
    cols_label(state = "State", grad_students = "Graduate Students") |>
    fmt_number(grad_students, decimals = 0) |>
    gt_color_rows(grad_students,
                  palette = "ggsci::blue_material",
                  direction = 1) |>
    tab_source_note("Source: IPEDS via NCES (nces.ed.gov/ipeds)")
```

```{r}
#| label: task6-table2

# Table 2: CUNY first-year undergraduates in 2024
cuny_fy <- IPEDS |>
    filter(is_cuny, year == 2024, level == "first year undergrad") |>
    group_by(institution_name) |>
    summarize(first_year = sum(enrollment_total), .groups = "drop") |>
    arrange(desc(first_year))

total_cuny_fy <- sum(cuny_fy$first_year)

cuny_fy |>
    mutate(pct = first_year / total_cuny_fy) |>
    gt() |>
    tab_header(
        title    = "CUNY First-Year Undergraduate Enrollment (2024)",
        subtitle = "Absolute enrollment and share of total CUNY first-year students"
    ) |>
    cols_label(
        institution_name = "Institution",
        first_year       = "First-Year Students",
        pct              = "% of CUNY Total"
    ) |>
    fmt_number(first_year, decimals = 0) |>
    fmt_percent(pct, decimals = 1) |>
    gt_color_rows(first_year,
                  palette = "ggsci::blue_material",
                  direction = 1) |>
    tab_source_note("Source: IPEDS via NCES")
```

```{r}
#| label: task6-table3

# Table 3: Baruch undergraduate enrollment with year-over-year % change
# lag() returns the previous row's value --- used here to compute YoY change
IPEDS |>
    filter(str_detect(institution_name, "Baruch"),
           level == "all undergrad") |>
    group_by(year) |>
    summarize(enrollment = sum(enrollment_total), .groups = "drop") |>
    arrange(year) |>
    mutate(
        prior_year = lag(enrollment),
        yoy_change = (enrollment - prior_year) / prior_year
    ) |>
    gt() |>
    tab_header(
        title    = "Baruch College Undergraduate Enrollment (2010-2024)",
        subtitle = "With year-over-year percent change"
    ) |>
    cols_label(
        year       = "Year",
        enrollment = "Enrollment",
        prior_year = "Prior Year",
        yoy_change = "YoY Change"
    ) |>
    fmt_number(c(enrollment, prior_year), decimals = 0) |>
    fmt_percent(yoy_change, decimals = 1) |>
    sub_missing(missing_text = "-") |>
    gt_highlight_rows(
        rows             = year == 2024,
        fill             = "#e8f4fd",
        bold_target_only = TRUE,
        target_col       = enrollment
    ) |>
    tab_source_note(
        "Source: IPEDS via NCES. 2024 row highlighted as first post-SFFA year."
    )
```

```{r}
#| label: task6-table4

# Table 4: Institutions where white student fraction fell most (2010 to 2020)
# Pre-filter to 1,000+ students to exclude small schools with noisy fluctuations
IPEDS |>
    filter(level == "all undergrad", year %in% c(2010, 2020)) |>
    mutate(white_total = enrollment_m_whit + enrollment_f_whit,
           white_frac  = white_total / enrollment_total) |>
    group_by(institution_id, institution_name, year) |>
    summarize(white_frac   = mean(white_frac),
              total_enroll = sum(enrollment_total),
              .groups = "drop") |>
    filter(total_enroll > 1000) |>
    select(institution_id, institution_name, year, white_frac) |>
    pivot_wider(names_from   = year,
                values_from  = white_frac,
                names_prefix = "frac_") |>
    mutate(change = frac_2020 - frac_2010) |>
    drop_na() |>
    slice_min(change, n = 5) |>
    select(institution_name, frac_2010, frac_2020, change) |>
    gt() |>
    tab_header(
        title    = "Largest Declines in White Student Share (2010-2020)",
        subtitle = "Among institutions enrolling 1,000+ undergraduates"
    ) |>
    cols_label(
        institution_name = "Institution",
        frac_2010        = "White % (2010)",
        frac_2020        = "White % (2020)",
        change           = "Change"
    ) |>
    fmt_percent(c(frac_2010, frac_2020, change), decimals = 1) |>
    tab_source_note("Source: IPEDS via NCES")
```

```{r}
#| label: task6-table5

# Table 5: States where female undergrad fraction increased most (2010 to 2024)
IPEDS |>
    filter(level == "all undergrad", year %in% c(2010, 2024)) |>
    group_by(state, year) |>
    summarize(female = sum(enrollment_f_total),
              total  = sum(enrollment_total),
              .groups = "drop") |>
    mutate(female_frac = female / total) |>
    select(state, year, female_frac) |>
    pivot_wider(names_from   = year,
                values_from  = female_frac,
                names_prefix = "frac_") |>
    mutate(change = frac_2024 - frac_2010) |>
    drop_na() |>
    slice_max(change, n = 3) |>
    gt() |>
    tab_header(
        title = "Top 3 States: Growth in Female Undergraduate Share (2010-2024)"
    ) |>
    cols_label(
        state     = "State",
        frac_2010 = "Female % (2010)",
        frac_2024 = "Female % (2024)",
        change    = "Change"
    ) |>
    fmt_percent(c(frac_2010, frac_2024, change), decimals = 1) |>
    tab_source_note("Source: IPEDS via NCES")
```

```{r}
#| label: task6-ec-kldiv
#| message: false

# Extra Credit: KL Divergence
# Finds the 3 institutions most representative of the national undergraduate
# racial composition. Lower KL divergence = closer to national average.
# Formula: D(P||Q) = sum(P * log(P/Q)) where P = national, Q = institution

# Step 1: Compute national undergraduate racial proportions in 2024
national_props <- IPEDS |>
    filter(level == "all undergrad", year == 2024) |>
    summarize(
        aian     = sum(enrollment_m_aian + enrollment_f_aian),
        asia     = sum(enrollment_m_asia + enrollment_f_asia),
        bkaa     = sum(enrollment_m_bkaa + enrollment_f_bkaa),
        hisp     = sum(enrollment_m_hisp + enrollment_f_hisp),
        nhpi     = sum(enrollment_m_nhpi + enrollment_f_nhpi),
        whit     = sum(enrollment_m_whit + enrollment_f_whit),
        two_more = sum(enrollment_m_2mor + enrollment_f_2mor),
        unkn     = sum(enrollment_m_unkn + enrollment_f_unkn),
        nral     = sum(enrollment_m_nral + enrollment_f_nral)
    ) |>
    mutate(total = aian + asia + bkaa + hisp + nhpi +
                   whit + two_more + unkn + nral) |>
    mutate(across(aian:nral, ~ . / total)) |>
    select(-total)

P <- unlist(national_props)

# Step 2: Compute each institution's proportions and KL divergence from P
kl_div <- IPEDS |>
    filter(level == "all undergrad", year == 2024) |>
    group_by(institution_id, institution_name) |>
    summarize(
        aian     = sum(enrollment_m_aian + enrollment_f_aian),
        asia     = sum(enrollment_m_asia + enrollment_f_asia),
        bkaa     = sum(enrollment_m_bkaa + enrollment_f_bkaa),
        hisp     = sum(enrollment_m_hisp + enrollment_f_hisp),
        nhpi     = sum(enrollment_m_nhpi + enrollment_f_nhpi),
        whit     = sum(enrollment_m_whit + enrollment_f_whit),
        two_more = sum(enrollment_m_2mor + enrollment_f_2mor),
        unkn     = sum(enrollment_m_unkn + enrollment_f_unkn),
        nral     = sum(enrollment_m_nral + enrollment_f_nral),
        total    = sum(enrollment_total),
        .groups  = "drop"
    ) |>
    filter(total >= 1000) |>
    mutate(
        q_aian     = aian     / total,
        q_asia     = asia     / total,
        q_bkaa     = bkaa     / total,
        q_hisp     = hisp     / total,
        q_nhpi     = nhpi     / total,
        q_whit     = whit     / total,
        q_two_more = two_more / total,
        q_unkn     = unkn     / total,
        q_nral     = nral     / total,
        # Nugget term 1e-10 prevents log(0) --- same approach as entropy
        kl = P[["aian"]]     * log(P[["aian"]]     / (q_aian     + 1e-10)) +
             P[["asia"]]     * log(P[["asia"]]     / (q_asia     + 1e-10)) +
             P[["bkaa"]]     * log(P[["bkaa"]]     / (q_bkaa     + 1e-10)) +
             P[["hisp"]]     * log(P[["hisp"]]     / (q_hisp     + 1e-10)) +
             P[["nhpi"]]     * log(P[["nhpi"]]     / (q_nhpi     + 1e-10)) +
             P[["whit"]]     * log(P[["whit"]]     / (q_whit     + 1e-10)) +
             P[["two_more"]] * log(P[["two_more"]] / (q_two_more + 1e-10)) +
             P[["unkn"]]     * log(P[["unkn"]]     / (q_unkn     + 1e-10)) +
             P[["nral"]]     * log(P[["nral"]]     / (q_nral     + 1e-10))
    ) |>
    slice_min(kl, n = 3) |>
    select(institution_name, kl, total)

kl_div |>
    gt() |>
    tab_header(
        title    = "Most Representative Institutions (2024)",
        subtitle = md("3 universities (1,000+ undergrads) whose racial composition
                      most closely mirrors the national average, by KL divergence")
    ) |>
    cols_label(
        institution_name = "Institution",
        kl               = "KL Divergence",
        total            = "Total Undergrads"
    ) |>
    fmt_number(kl, decimals = 4) |>
    fmt_number(total, decimals = 0) |>
    tab_source_note(
        md("KL divergence measures dissimilarity from the national undergraduate
        racial composition. A value of zero would indicate a perfect match.
        Source: IPEDS via NCES")
    )
```

Taken together, these tables reveal a higher education system in slow but
steady demographic transition --- one that predates SFFA by at least a decade.
The declining white student share at several large institutions, the growing
female majority in several states, and the consistent diversity of CUNY's
entering classes all point to structural shifts driven by changing national
demographics rather than any single policy event. This context is essential
for interpreting the post-SFFA data in the section that follows.

## Final Insights and Deliverable

### Op-Ed Supporting Calculations

```{r}
#| label: oped-calculations
#| message: false

# Diversity metric 1: fraction of non-White, non-Asian undergraduates
# Diversity metric 2: Shannon entropy (extra credit)

# --- Baruch first-year class: 2022, 2023, 2024 ---
baruch_fy <- IPEDS |>
    filter(str_detect(institution_name, "Baruch"),
           level == "first year undergrad",
           year %in% c(2022, 2023, 2024)) |>
    mutate(
        minority = enrollment_m_bkaa + enrollment_f_bkaa +
                   enrollment_m_hisp + enrollment_f_hisp +
                   enrollment_m_aian + enrollment_f_aian +
                   enrollment_m_nhpi + enrollment_f_nhpi +
                   enrollment_m_2mor + enrollment_f_2mor
    ) |>
    group_by(year) |>
    summarize(
        total     = sum(enrollment_total),
        diversity = sum(minority) / sum(enrollment_total),
        .groups   = "drop"
    )

baruch_div_2022      <- baruch_fy |> filter(year == 2022) |> pull(diversity)
baruch_div_2023      <- baruch_fy |> filter(year == 2023) |> pull(diversity)
baruch_div_2024      <- baruch_fy |> filter(year == 2024) |> pull(diversity)
baruch_total_fy_2024 <- baruch_fy |> filter(year == 2024) |> pull(total)
baruch_change        <- baruch_div_2024 - baruch_div_2023

# --- Long-term trend: 2010 vs 2024 ---
baruch_longterm <- IPEDS |>
    filter(str_detect(institution_name, "Baruch"),
           level == "first year undergrad",
           year %in% c(2010, 2024)) |>
    mutate(
        minority = enrollment_m_bkaa + enrollment_f_bkaa +
                   enrollment_m_hisp + enrollment_f_hisp +
                   enrollment_m_aian + enrollment_f_aian +
                   enrollment_m_nhpi + enrollment_f_nhpi +
                   enrollment_m_2mor + enrollment_f_2mor
    ) |>
    group_by(year) |>
    summarize(
        diversity = sum(minority) / sum(enrollment_total),
        .groups   = "drop"
    )

baruch_div_2010        <- baruch_longterm |> filter(year == 2010) |> pull(diversity)
baruch_longterm_change <- baruch_div_2024 - baruch_div_2010

# --- Whole student body YoY (all undergrads, not just first-year) ---
baruch_allug <- IPEDS |>
    filter(str_detect(institution_name, "Baruch"),
           level == "all undergrad",
           year %in% c(2023, 2024)) |>
    mutate(
        minority = enrollment_m_bkaa + enrollment_f_bkaa +
                   enrollment_m_hisp + enrollment_f_hisp +
                   enrollment_m_aian + enrollment_f_aian +
                   enrollment_m_nhpi + enrollment_f_nhpi +
                   enrollment_m_2mor + enrollment_f_2mor
    ) |>
    group_by(year) |>
    summarize(
        diversity = sum(minority) / sum(enrollment_total),
        .groups   = "drop"
    )

baruch_allug_2023   <- baruch_allug |> filter(year == 2023) |> pull(diversity)
baruch_allug_2024   <- baruch_allug |> filter(year == 2024) |> pull(diversity)
baruch_allug_change <- baruch_allug_2024 - baruch_allug_2023

# --- UC Berkeley comparison ---
cal_fy <- IPEDS |>
    filter(str_detect(institution_name, "University of California-Berkeley"),
           level == "first year undergrad",
           year %in% c(2022, 2023, 2024)) |>
    mutate(
        minority = enrollment_m_bkaa + enrollment_f_bkaa +
                   enrollment_m_hisp + enrollment_f_hisp +
                   enrollment_m_aian + enrollment_f_aian +
                   enrollment_m_nhpi + enrollment_f_nhpi +
                   enrollment_m_2mor + enrollment_f_2mor
    ) |>
    group_by(year) |>
    summarize(
        total     = sum(enrollment_total),
        diversity = sum(minority) / sum(enrollment_total),
        .groups   = "drop"
    )

cal_div_2023 <- cal_fy |> filter(year == 2023) |> pull(diversity)
cal_div_2024 <- cal_fy |> filter(year == 2024) |> pull(diversity)
cal_change   <- cal_div_2024 - cal_div_2023

# --- Cal State LA comparison (urban commuter, more comparable to Baruch) ---
calstate_fy <- IPEDS |>
    filter(str_detect(institution_name, "California State University-Los Angeles"),
           level == "first year undergrad",
           year %in% c(2022, 2023, 2024)) |>
    mutate(
        minority = enrollment_m_bkaa + enrollment_f_bkaa +
                   enrollment_m_hisp + enrollment_f_hisp +
                   enrollment_m_aian + enrollment_f_aian +
                   enrollment_m_nhpi + enrollment_f_nhpi +
                   enrollment_m_2mor + enrollment_f_2mor
    ) |>
    group_by(year) |>
    summarize(
        total     = sum(enrollment_total),
        diversity = sum(minority) / sum(enrollment_total),
        .groups   = "drop"
    )

calstate_div_2023 <- calstate_fy |> filter(year == 2023) |> pull(diversity)
calstate_div_2024 <- calstate_fy |> filter(year == 2024) |> pull(diversity)
calstate_change   <- calstate_div_2024 - calstate_div_2023
```

```{r}
#| label: entropy-calculations
#| message: false

# Extra Credit: Shannon entropy diversity metric
# H = -sum(p * log(p)) across all racial groups
# Higher entropy = more evenly distributed = more diverse
# Nugget term 1e-10 handles p=0 without producing NA

compute_entropy <- function(enrollment_row) {
    groups <- c(
        enrollment_row[["enrollment_m_aian"]] + enrollment_row[["enrollment_f_aian"]],
        enrollment_row[["enrollment_m_asia"]] + enrollment_row[["enrollment_f_asia"]],
        enrollment_row[["enrollment_m_bkaa"]] + enrollment_row[["enrollment_f_bkaa"]],
        enrollment_row[["enrollment_m_hisp"]] + enrollment_row[["enrollment_f_hisp"]],
        enrollment_row[["enrollment_m_nhpi"]] + enrollment_row[["enrollment_f_nhpi"]],
        enrollment_row[["enrollment_m_whit"]] + enrollment_row[["enrollment_f_whit"]],
        enrollment_row[["enrollment_m_2mor"]] + enrollment_row[["enrollment_f_2mor"]]
    )
    total <- sum(groups)
    if (total == 0) return(NA_real_)
    props <- groups / total
    -sum(props * log(props + 1e-10))
}

# Baruch entropy by year
baruch_entropy <- IPEDS |>
    filter(str_detect(institution_name, "Baruch"),
           level == "first year undergrad",
           year %in% c(2022, 2023, 2024)) |>
    rowwise() |>
    mutate(entropy = compute_entropy(cur_data())) |>
    ungroup() |>
    group_by(year) |>
    summarize(avg_entropy = mean(entropy, na.rm = TRUE), .groups = "drop")

baruch_entropy_2022   <- baruch_entropy |> filter(year == 2022) |> pull(avg_entropy)
baruch_entropy_2023   <- baruch_entropy |> filter(year == 2023) |> pull(avg_entropy)
baruch_entropy_2024   <- baruch_entropy |> filter(year == 2024) |> pull(avg_entropy)
baruch_entropy_change <- baruch_entropy_2024 - baruch_entropy_2023

# UC Berkeley entropy
cal_entropy <- IPEDS |>
    filter(str_detect(institution_name, "University of California-Berkeley"),
           level == "first year undergrad",
           year %in% c(2022, 2023, 2024)) |>
    rowwise() |>
    mutate(entropy = compute_entropy(cur_data())) |>
    ungroup() |>
    group_by(year) |>
    summarize(avg_entropy = mean(entropy, na.rm = TRUE), .groups = "drop")

cal_entropy_2023   <- cal_entropy |> filter(year == 2023) |> pull(avg_entropy)
cal_entropy_2024   <- cal_entropy |> filter(year == 2024) |> pull(avg_entropy)
cal_entropy_change <- cal_entropy_2024 - cal_entropy_2023
```

```{r}
#| label: oped-viz
#| message: false
#| fig-cap: "Diversity trends in first-year undergraduate classes at Baruch College vs UC Berkeley (2010-2024), measured by Shannon entropy. Dashed line marks the SFFA ruling."
#| fig-alt: "Line chart showing Shannon entropy of racial enrollment at Baruch College and UC Berkeley from 2010 to 2024, with a dashed vertical line at 2023 marking the SFFA ruling."

library(ggplot2)

entropy_trend <- IPEDS |>
    filter(
        (str_detect(institution_name, "Baruch") |
         str_detect(institution_name, "University of California-Berkeley")),
        level == "first year undergrad"
    ) |>
    mutate(
        total_counted = enrollment_m_aian + enrollment_f_aian +
                        enrollment_m_asia + enrollment_f_asia +
                        enrollment_m_bkaa + enrollment_f_bkaa +
                        enrollment_m_hisp + enrollment_f_hisp +
                        enrollment_m_nhpi + enrollment_f_nhpi +
                        enrollment_m_whit + enrollment_f_whit +
                        enrollment_m_2mor + enrollment_f_2mor,
        p_aian = (enrollment_m_aian + enrollment_f_aian) / total_counted,
        p_asia = (enrollment_m_asia + enrollment_f_asia) / total_counted,
        p_bkaa = (enrollment_m_bkaa + enrollment_f_bkaa) / total_counted,
        p_hisp = (enrollment_m_hisp + enrollment_f_hisp) / total_counted,
        p_nhpi = (enrollment_m_nhpi + enrollment_f_nhpi) / total_counted,
        p_whit = (enrollment_m_whit + enrollment_f_whit) / total_counted,
        p_2mor = (enrollment_m_2mor + enrollment_f_2mor) / total_counted,
        entropy = -(p_aian * log(p_aian + 1e-10) +
                    p_asia * log(p_asia + 1e-10) +
                    p_bkaa * log(p_bkaa + 1e-10) +
                    p_hisp * log(p_hisp + 1e-10) +
                    p_nhpi * log(p_nhpi + 1e-10) +
                    p_whit * log(p_whit + 1e-10) +
                    p_2mor * log(p_2mor + 1e-10)),
        institution_label = if_else(
            str_detect(institution_name, "Baruch"),
            "Baruch College (CUNY)",
            "UC Berkeley (CA baseline)"
        )
    ) |>
    group_by(institution_label, year) |>
    summarize(entropy = mean(entropy, na.rm = TRUE), .groups = "drop")

ggplot(entropy_trend,
       aes(x = year, y = entropy,
           color = institution_label,
           group = institution_label)) +
    geom_line(linewidth = 1.2) +
    geom_point(size = 2.5) +
    geom_vline(xintercept = 2023, linetype = "dashed",
               color = "grey40", linewidth = 0.8) +
    annotate("text", x = 2023.15,
             y = min(entropy_trend$entropy, na.rm = TRUE),
             label = "SFFA ruling", hjust = 0,
             size = 3.2, color = "grey40") +
    scale_color_manual(
        values = c("Baruch College (CUNY)"     = "#003865",
                   "UC Berkeley (CA baseline)" = "#FDB515")
    ) +
    scale_x_continuous(breaks = seq(2010, 2024, 2)) +
    labs(
        title    = "Diversity Trends in First-Year Undergraduate Classes",
        subtitle = "Shannon entropy of racial composition, 2010-2024",
        x        = "Year",
        y        = "Entropy (higher = more diverse)",
        color    = NULL,
        caption  = "Source: IPEDS via NCES. Dashed line marks SFFA ruling (June 2023)."
    ) +
    theme_minimal(base_size = 13) +
    theme(
        legend.position  = "bottom",
        plot.title       = element_text(face = "bold"),
        panel.grid.minor = element_blank()
    )
```

---

### Op-Ed

*The following Op-Ed is written from the perspective of the President of
Baruch College for publication in the campus newspaper.*

---

**One Year After SFFA: What Our Numbers Actually Show**  
*By the President of Baruch College*

Last June marked one year since the Supreme Court's ruling in *Students for
Fair Admissions v. Harvard* reshaped college admissions across the country.
Commentators predicted sweeping demographic shifts. Now that the data is in,
it's time to share what we're actually seeing at Baruch --- and what it means.

For this analysis, I measure diversity using Shannon entropy --- a statistical
measure borrowed from information theory that captures how evenly enrollment
is distributed across *all* racial groups simultaneously, rather than
collapsing everything into a single "percent minority" figure. A higher
entropy means a more genuinely balanced student body; a lower entropy means
enrollment is concentrated in fewer groups. By this measure, Baruch's
entering class had an entropy of `r round(baruch_entropy_2024, 3)` in Fall
2024, compared to `r round(baruch_entropy_2023, 3)` the year before and
`r round(baruch_entropy_2022, 3)` in 2022. The year-over-year shift was
`r round(abs(baruch_entropy_change), 3)`
`r if_else(baruch_entropy_change >= 0, "upward", "downward")`. Using the
simpler "percent of color" metric, our incoming class of
`r comma(baruch_total_fy_2024)` first-year students was
**`r percent(baruch_div_2024, accuracy=0.1)`** students of color in Fall 2024,
compared to `r percent(baruch_div_2023, accuracy=0.1)` the year before and
`r percent(baruch_div_2022, accuracy=0.1)` in 2022.

Looking beyond the entering class to Baruch's entire undergraduate student
body tells a similar story. Across all enrolled undergraduates,
`r percent(baruch_allug_2024, accuracy=0.1)` identified as students of color
in Fall 2024, compared to `r percent(baruch_allug_2023, accuracy=0.1)` the
prior year --- a shift of `r percent(abs(baruch_allug_change), accuracy=0.1)`
`r if_else(baruch_allug_change >= 0, "upward", "downward")`. Because the full
student body reflects multiple entering cohorts, it changes more slowly than
the entering class alone --- making even small year-over-year shifts meaningful
signals.

To put this in context, consider California's public universities ---
institutions that have operated without race-conscious admissions since
Proposition 209 passed in 1996, nearly three decades before SFFA. They
represent our best available baseline for what "normal" year-to-year
fluctuation looks like in the complete absence of race-conscious policy.

At UC Berkeley, first-year class diversity changed by
`r percent(abs(cal_change), accuracy=0.1)`
`r if_else(cal_change >= 0, "upward", "downward")` between 2023 and 2024.
At Cal State LA --- a large urban commuter school whose mission more closely
resembles Baruch's than Berkeley's does --- the shift was
`r percent(abs(calstate_change), accuracy=0.1)`
`r if_else(calstate_change >= 0, "upward", "downward")`.
Baruch's change of `r percent(abs(baruch_change), accuracy=0.1)` sits
`r if_else(abs(baruch_change) > abs(cal_change), "above", "within")`
that California range --- suggesting our shift
`r if_else(abs(baruch_change) > abs(cal_change), "may reflect a genuine SFFA effect on our admissions pipeline.", "is consistent with the normal variation seen even at schools entirely unaffected by the ruling.")`

Zooming out further: since 2010, Baruch's first-year diversity has shifted
by `r percent(abs(baruch_longterm_change), accuracy=0.1)`
`r if_else(baruch_longterm_change >= 0, "upward", "downward")` in total ---
a long-term trend that dwarfs any single-year post-SFFA movement and
reflects the changing demographics of New York City itself rather than
any admissions policy.

These numbers tell only part of the story. Baruch has always derived its
strength from serving New York City --- one of the most diverse cities on
Earth. Our students come from over 150 countries, speak dozens of languages,
and represent every borough. That diversity is written into our mission and
has not changed because of a Supreme Court ruling.

We will continue to pursue diversity through every permissible avenue:
outreach to underserved communities, partnerships with NYC high schools,
robust financial aid, and a campus culture that actively welcomes
first-generation college students. The Court told us *how* we can pursue
diversity --- not *whether* we should.

The data will continue to evolve as more post-SFFA admissions cycles
complete. We are committed to transparency, and will share our numbers
every year. Baruch's doors remain open --- and our commitment to reflecting
the full breadth of New York City remains unchanged.

---

::: {.callout-note title="AI Usage Statement" collapse="true"}
Claude (Anthropic) was used to help structure and write portions of the
code in this project and to suggest a document outline. All non-code text,
including the Op-Ed, was written by the student. No AI tools were used to
generate or edit prose directly.
:::

------------------------------------------------------------------------

This work ©2025 by dhwanikan was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at
[MP #01](https://michael-weylandt.com/STA9750/mini/mini01.html)
