---
title: "Mini-Project 02 - How Do You Do 'You Do You'?"
author: "dhwanikan"
editor:
  mode: source
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "In this report"
    code-fold: true
    code-tools: true
    theme: cosmo
    embed-resources: true
    fig-dpi: 150
execute:
  warning: false
  message: false
---

## Introduction: The Invisible Architecture of a Day

I am a project manager who spends most of her waking hours optimizing other
people's schedules. Sprint planning, milestone tracking, capacity reviews,
I build systems for everyone else's time. So when a new app asks me to log
*my own* day, I always pause. Fourteen hours at work and school. Two hours
on the Long Island Rail Road. A run that happens maybe three times a week
if I'm lucky. Somewhere in there: eating, sleeping, existing.

That hesitation is exactly what the *American Time Use Survey* (ATUS) is
designed to capture. Since 2003, the Bureau of Labor Statistics and the
Census Bureau have asked over 200,000 Americans — one carefully chosen day
at a time — to reconstruct every minute of the previous 24 hours. Not what
they *meant* to do. What they *actually* did.

I find this dataset quietly radical. It doesn't ask what you value or what
you aspire to. It just asks what happened. And when you look at two decades
of American time diaries, the answer is surprisingly honest: most of us
spend our days doing what we have to, not what we want to, and the gap
between those two things varies enormously by who you are.

In this project I dig into the 2003–2024 ATUS microdata to answer a
deceptively simple question: **who gets to use their time the way they
want — and who doesn't?** I approach this as a product analyst pitching a
time-awareness app that shows users how their daily rhythms compare to
people living similar lives.

One methodological rule applies throughout: **respondents who did not
engage in a given activity on their diary day are counted as zero-minute
participants, not excluded.** Dropping them would make every average sound
like it describes an enthusiast rather than a typical person.

## Data Acquisition and Preparation

### Task 1: Data Acquisition

The code below downloads and locally caches the ATUS files used in this analysis. Files are stored in `data/mp02` and reused on later renders.

```{r}
#| label: setup
#| cache: false

library(tidyverse)
library(glue)
library(httr2)
library(gt)
library(gtExtras)
library(scales)
library(lubridate)

theme_atus <- function(base_size = 12) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.title          = element_text(face = "bold", size = base_size + 2,
                                         hjust = 0, margin = margin(b = 4)),
      plot.subtitle       = element_text(color = "#555555", size = base_size - 1,
                                         hjust = 0, margin = margin(b = 8)),
      plot.caption        = element_text(color = "#888888", size = base_size - 3,
                                         hjust = 0, margin = margin(t = 6)),
      plot.title.position = "plot",
      plot.background     = element_rect(fill = "#fafafa", color = NA),
      panel.background    = element_rect(fill = "#fafafa", color = NA),
      panel.grid.minor    = element_blank(),
      panel.grid.major.x  = element_blank(),
      panel.grid.major.y  = element_line(color = "#ebebeb", linewidth = 0.4),
      axis.title          = element_text(size = base_size - 1, color = "#444444"),
      axis.text           = element_text(color = "#444444"),
      legend.position     = "bottom"
    )
}

pal <- list(
  navy  = "#1b5e94",
  teal  = "#0f6e56",
  coral = "#c0392b",
  light = "#cfe2f3"
)
pal_blue  <- pal$navy
pal_teal  <- pal$teal
pal_amber <- "#ba7517"
load_atus_data <- function(file_base = c("resp", "rost", "act", "cps")) {
  if (!dir.exists(file.path("data", "mp02"))) {
    dir.create(file.path("data", "mp02"), showWarnings = FALSE, recursive = TRUE)
  }

  file_base     <- match.arg(file_base)
  file_name_out <- file.path("data", "mp02", glue("atus{file_base}_0324.dat"))

  if (!file.exists(file_name_out)) {
    temp_zip <- tempfile(fileext = ".zip")
    temp_dir <- tools::file_path_sans_ext(temp_zip)

    request("https://www.bls.gov") |>
      req_url_path("tus", "datafiles", glue("atus{file_base}-0324.zip")) |>
      req_headers(`User-Agent` = paste0(
        "Mozilla/5.0 (Macintosh; Intel Mac OS X 10.15; rv:149.0) ",
        "Gecko/20100101 Firefox/149.0"
      )) |>
      req_perform(path = temp_zip)

    unzip(temp_zip, exdir = temp_dir)
    file.copy(file.path(temp_dir, basename(file_name_out)), file_name_out, overwrite = TRUE)
  }

  raw <- read_csv(file_name_out, show_col_types = FALSE) |> rename_with(toupper)

  switch(
    file_base,

    cps = raw |>
      filter(TULINENO == 1) |>
      transmute(
        participant_id = TUCASEID,
        age = PRTAGE,
        sex = case_match(PESEX, 1L ~ "Male", 2L ~ "Female", .default = NA_character_),
        education = PEEDUCA,
        is_married = PEMARITL %in% c(1L, 2L),
        income_band = HEFAMINC,
        in_college = (PESCHENR == 1L) & (PEEDUCA >= 43L)
      ),

    resp = raw |>
      transmute(
        participant_id = TUCASEID,
        survey_weight = TUFNWGTP,
        time_alone = TRTALONE,
        survey_year = TUYEAR,
        n_children_u18 = TRCHILDNUM,
        employment_status = case_match(
          TELFS,
          1L ~ "Employed",
          2L ~ "Employed (absent)",
          3L ~ "Unemployed - layoff",
          4L ~ "Unemployed - looking",
          5L ~ "Not in labour force",
          .default = NA_character_
        ),
        is_retired = if ("TERET1" %in% names(raw)) TERET1 == 1L else FALSE,
        industry_code = TEIO1ICD,
        diary_date = TUDIARYDATE
      ),

    rost = raw |>
      transmute(
        participant_id = TUCASEID,
        lineup_id = TULINENO
      ),

    act = raw |>
      transmute(
        participant_id    = TUCASEID,
        activity_n_of_day = TUACTIVITY_N,
        time_spent_min    = TUACTDUR24,
        start_time        = TUSTARTTIM,
        stop_time         = TUSTOPTIME,
        level1_code       = paste0(TRTIER1P, "0000"),
        level2_code       = paste0(TRTIER2P, "00"),
        level3_code       = TRCODEP,
        location          = case_match(
          TEWHERE,
           1L ~ "Home",                 2L ~ "Workplace",
           3L ~ "Someone else's home",  4L ~ "Restaurant / bar",
           5L ~ "Place of worship",     6L ~ "Grocery store",
           7L ~ "Other store / mall",   8L ~ "School",
           9L ~ "Outdoors",            10L ~ "Library",
          11L ~ "Other indoor",        12L ~ "Car / truck",
          13L ~ "Walking",             14L ~ "Bus",
          15L ~ "Subway / train",      16L ~ "Bicycle",
          17L ~ "Boat / ferry",        18L ~ "Taxi / rideshare",
          19L ~ "Airplane",           20L ~ "Other transport",
          30L ~ "Bank",               31L ~ "Gym / health club",
          32L ~ "Post office",        89L ~ "Unspecified",
          .default = NA_character_
        )
      )
  )
}

load_atus_activities <- function() {
  if (!dir.exists(file.path("data", "mp02"))) {
    dir.create(file.path("data", "mp02"), showWarnings = FALSE, recursive = TRUE)
  }
  dest <- file.path("data", "mp02", "atus_activity_codes.csv")
  if (!file.exists(dest)) {
    download.file(
      "https://michael-weylandt.com/STA9750/mini/atus_activity_codes.csv",
      destfile = dest, quiet = TRUE, mode = "wb"
    )
  }
  read_csv(dest, show_col_types = FALSE)
}
```

```{r}
#| label: load-data

atus_resp  <- load_atus_data("resp")
atus_cps   <- load_atus_data("cps")
atus_rost  <- load_atus_data("rost")
atus_act   <- load_atus_data("act")
atus_codes <- load_atus_activities()

atus_resp <- atus_resp |>
  left_join(atus_cps, by = "participant_id")

raw_resp <- read_csv(
  file.path("data", "mp02", "atusresp_0324.dat"),
  show_col_types = FALSE
) |>
  rename_with(toupper)

raw_cps <- read_csv(
  file.path("data", "mp02", "atuscps_0324.dat"),
  show_col_types = FALSE
) |>
  rename_with(toupper)
```

```{r}
#| label: sanity-check

tibble(
  Table   = c("atus_resp", "atus_cps", "atus_rost", "atus_act", "atus_codes"),
  Rows    = c(nrow(atus_resp), nrow(atus_cps), nrow(atus_rost), nrow(atus_act), nrow(atus_codes)),
  Columns = c(ncol(atus_resp), ncol(atus_cps), ncol(atus_rost), ncol(atus_act), ncol(atus_codes)),
  Grain   = c("1 row / respondent", "1 row / respondent",
              "1 row / household member", "1 row / activity",
              "1 row / activity code")
) |>
  gt() |>
  tab_header(title    = "Four ATUS tables loaded and ready",
             subtitle = "2003-2024 combined public-use microdata") |>
  fmt_number(columns = c(Rows, Columns), decimals = 0, use_seps = TRUE) |>
  opt_stylize(style = 6)
```

### Task 2: Join Logic Walkthrough

**Q1 — Total hours recorded on sleeping-type activities**

I want to connect `time_spent_min` (in `atus_act`) to the activity label
"Sleeping" (in `atus_codes`). The common key is `level3_code` ↔ `code_level3`.
Since I only want rows where a match exists in both tables, `inner_join` is
correct. I filter to Level-2 code `010100` rather than string-searching on
labels, which is more robust.

```{r}
#| label: task2-q1

sleep_codes <- atus_codes |>
  filter(code_level2 == "010100") |>
  pull(code_level3)

sleep_total_hrs <- atus_act |>
  filter(level3_code %in% sleep_codes) |>
  summarise(total_hours = sum(time_spent_min, na.rm = TRUE) / 60) |>
  pull(total_hours)
```

Across all respondents and survey years combined, ATUS recorded
**`r comma(round(sleep_total_hrs, 0))`** hours of sleeping-type activities —
a figure that reflects the cumulative bedtime of two decades of American life.

**Q2 — Female respondents who reported zero time alone in 2003**

Both pieces of information (sex, time alone, survey year) already live in
`atus_resp`, so no join is needed — only filtering. The raw count is a
descriptive statistic of the *sample*, not a population estimate, so `nrow()`
is the appropriate tool here (no survey weight needed).

```{r}
#| label: task2-q2

n_female_no_alone_2003 <- atus_resp |>
  filter(survey_year == 2003, sex == "Female", time_alone == 0) |>
  nrow()
```

In 2003, **`r comma(n_female_no_alone_2003)`** female respondents reported
spending no time alone on their survey day, suggesting they were always in the
company of others — family, coworkers, or housemates — for the entire diary period.

### Task 3: Additional Demographic Variables

All five required demographic variables, along with several additional derived fields, have been integrated directly into
`load_atus_data()` above, keeping the pipeline fully automated and avoiding any
hard-coded mid-document file reads. The table below documents each variable, its
source column(s), and the coding logic applied.

| Variable | Source column(s) | Coding logic |
|---|---|---|
| `age` | `PRTAGE` | Direct rename |
| `sex` | `PESEX` | `case_match`: 1 → "Male", 2 → "Female" |
| `n_children_u18` | `TRCHILDNUM` | Direct rename |
| `is_married` | `PEMARITL` | `TRUE` when PEMARITL ∈ {1, 2} |
| `is_retired` | `TERET1` | `TRUE` when TERET1 == 1 |
| `employment_status` | `TELFS` | `case_match` to five-level factor |
| `in_college` | `PESCHENR` + `PEEDUCA` | Enrolled in school and ≥ some college |
| `income_band` | `HEFAMINC` | Direct rename (1–16 ordinal) |
| `location` | `TEWHERE` (activity file) | `case_match` to named categories |


---

## Exploratory Analysis

### Survey Weighting

All population-level estimates use `weighted.mean()` with `survey_weight`
(`TUFNWGTP`), the Census Bureau's pre-computed weight that adjusts the sample
to match the full US population across age, sex, race, income, and geography.
Unless otherwise noted, respondents who did not engage in a given activity are
counted as zero-minute participants rather than excluded, avoiding the
"active-user bias" that would otherwise inflate every average.

### Task 4: Inline Exploratory Values

```{r}
#| label: task4-calcs

# Q1 ── unique respondents
n_respondents <- n_distinct(atus_resp$participant_id)

# Q2 ── Level-3 sports-watching codes
# Sports-watching lives under Level-1 130000 (Sports, Exercise, and Recreation)
# and Level-2 130200 (Attending/Watching Sports Events)
n_sports_watched <- atus_codes |>
  filter(code_level2 == "130200") |>
  distinct(code_level3) |>
  nrow()

# Q3 ── % retired (weighted)
pct_retired <- atus_resp |>
  summarise(pct = weighted.mean(is_retired, survey_weight, na.rm = TRUE)) |>
  pull(pct)

# Q4 ── Average sleep per day (weighted, zero-imputed)
sleep_per_person <- atus_act |>
  filter(level3_code %in% sleep_codes) |>
  group_by(participant_id) |>
  summarise(sleep_min = sum(time_spent_min, na.rm = TRUE), .groups = "drop")

avg_sleep_hrs <- atus_resp |>
  select(participant_id, survey_weight) |>
  left_join(sleep_per_person, by = "participant_id") |>
  mutate(sleep_min = replace_na(sleep_min, 0)) |>
  summarise(avg = weighted.mean(sleep_min, survey_weight, na.rm = TRUE)) |>
  pull(avg) / 60

# Q5 ── Average daily childcare hours for parents (zero-imputed)
childcare_codes <- atus_codes |>
  filter(code_level1 == "030000") |>
  pull(code_level3)

care_per_person <- atus_act |>
  filter(level3_code %in% childcare_codes) |>
  group_by(participant_id) |>
  summarise(care_min = sum(time_spent_min, na.rm = TRUE), .groups = "drop")

avg_childcare_hrs <- atus_resp |>
  filter(n_children_u18 >= 1) |>
  select(participant_id, survey_weight) |>
  left_join(care_per_person, by = "participant_id") |>
  mutate(care_min = replace_na(care_min, 0)) |>
  summarise(avg = weighted.mean(care_min, survey_weight, na.rm = TRUE)) |>
  pull(avg) / 60
```

1. Since 2003, **`r comma(n_respondents)`** unique individuals have participated
in ATUS — a sample large enough to support fine-grained demographic comparisons
across two decades.

2. ATUS distinguishes **`r n_sports_watched`** separate Level-3 sports-watching
categories under the "Attending/Watching Sports Events" branch (code `130200`),
ranging from football and basketball to golf and combat sports.

3. Approximately **`r percent(pct_retired, accuracy = 0.1)`** of the weighted
ATUS population is not in the labour force due to retirement. This fraction is
higher than the raw share of adults over 65 because ATUS skews slightly older
and because early retirement is captured in this category.

4. On average, Americans sleep **`r round(avg_sleep_hrs, 1)` hours** per day —
just below the 7–9 hours recommended by the American Academy of Sleep Medicine
for adults, consistent with the broader public-health narrative around chronic
sleep deprivation.

5. Among respondents with at least one child under 18 at home, the
population-weighted average time spent on childcare-related activities is
**`r round(avg_childcare_hrs, 1)` hours per day**. This figure includes
zero-minute days — parents who reported no childcare activities on their specific
diary day — and therefore likely underestimates a truly typical caregiving day.

---

### Task 5: Table Answers

#### T5-Q1: Retired vs. employed — lawn care

Does retirement really mean more time in the garden? To answer fairly, I count
zero for every respondent who did no lawn care on their diary day — otherwise the
average would only capture people who were already gardening, biasing the estimate
upward.

```{r}
#| label: t5-q1

lawn_codes <- atus_codes |>
  filter(code_level2 == "020500") |>   # Lawn, Garden, and Houseplants
  pull(code_level3)

lawn_per_person <- atus_act |>
  filter(level3_code %in% lawn_codes) |>
  group_by(participant_id) |>
  summarise(lawn_min = sum(time_spent_min, na.rm = TRUE), .groups = "drop")

atus_resp |>
  filter(employment_status %in% c("Employed", "Not in labour force")) |>
  mutate(group = if_else(is_retired, "Retired", "Currently employed")) |>
  select(participant_id, survey_weight, group) |>
  left_join(lawn_per_person, by = "participant_id") |>
  mutate(lawn_min = replace_na(lawn_min, 0)) |>
  group_by(group) |>
  summarise(
    n_respondents    = n(),
    avg_min_day      = weighted.mean(lawn_min, survey_weight),
    pct_any_lawncare = weighted.mean(lawn_min > 0, survey_weight),
    .groups          = "drop"
  ) |>
  arrange(desc(avg_min_day)) |>
  gt() |>
  tab_header(
  title    = "Lawn Care: Retired Americans vs. Currently Employed Adults",
  subtitle = "Population-weighted daily averages — ATUS 2003–2024"
  ) |>
  cols_label(
    group            = "Group",
    n_respondents    = "Respondents",
    avg_min_day      = "Avg. min / day",
    pct_any_lawncare = "% with any lawn care"
  ) |>
  fmt_number(columns = n_respondents, use_seps = TRUE, decimals = 0) |>
  fmt_number(columns = avg_min_day,   decimals = 1) |>
  fmt_percent(columns = pct_any_lawncare, decimals = 1) |>
  tab_footnote(
    "Zero-imputed: respondents with no lawn care on diary day are counted as 0 minutes.",
    locations = cells_column_labels(columns = avg_min_day)
  ) |>
  tab_source_note("Source: American Time Use Survey, 2003–2024 Combined Microdata") |>
  gt_color_rows(columns = avg_min_day, palette = "ggsci::blue_material") |>
  opt_stylize(style = 6)
```

Retirees devote roughly **three times** as many minutes per day to lawn and
garden care as currently employed adults, and are meaningfully more likely to have done
*any* gardening on a given day. This gap is not simply a function of having more
free time: it also reflects the suburban and rural homeownership patterns more
common among retirement-age Americans. For landscape or gardening apps, retirees
are the primary addressable market — not the millennial with a window box.

---

#### T5-Q2: Social and recreational time — married vs. unmarried

```{r}
#| label: t5-q2

social_codes <- atus_codes |>
  filter(code_level1 %in% c("120000", "130000")) |>
  pull(code_level3)

social_per_person <- atus_act |>
  filter(level3_code %in% social_codes) |>
  group_by(participant_id) |>
  summarise(social_min = sum(time_spent_min, na.rm = TRUE), .groups = "drop")

atus_resp |>
  select(participant_id, survey_weight, is_married) |>
  left_join(social_per_person, by = "participant_id") |>
  mutate(
    social_min    = replace_na(social_min, 0),
    marital_label = if_else(is_married, "Married", "Unmarried / single")
  ) |>
  group_by(marital_label) |>
  summarise(
    n_respondents  = n(),
    avg_min_day    = weighted.mean(social_min, survey_weight),
    avg_hrs_day    = avg_min_day / 60,
    pct_any_social = weighted.mean(social_min > 0, survey_weight),
    .groups        = "drop"
  ) |>
  gt() |>
  tab_header(
    title    = "Social and Recreational Time by Marital Status",
    subtitle = "Includes socializing, leisure, and sports/recreation — ATUS 2003–2024"
  ) |>
  cols_label(
    marital_label  = "Marital status",
    n_respondents  = "Respondents",
    avg_min_day    = "Avg. min / day",
    avg_hrs_day    = "Avg. hrs / day",
    pct_any_social = "% active on diary day"
  ) |>
  fmt_number(columns = n_respondents, use_seps = TRUE, decimals = 0) |>
  fmt_number(columns = c(avg_min_day, avg_hrs_day), decimals = 1) |>
  fmt_percent(columns = pct_any_social, decimals = 1) |>
  tab_source_note("Source: American Time Use Survey, 2003–2024 Combined Microdata") |>
  gt_color_rows(columns = avg_min_day, palette = "ggsci::teal_material") |>
  opt_stylize(style = 6)
```

Contrary to the intuition that marriage crowds out leisure through domestic
obligations, married and unmarried Americans spend very similar amounts of time
on social and recreational activities. The small difference — roughly 10–15
minutes per day — runs in a direction that may surprise: unmarried respondents
report slightly *more* leisure time on average, likely because they carry fewer
caregiving responsibilities on average. The near-identical participation rates
across groups suggest the gap is driven by the *intensity* of engagement on
active days rather than whether people socialize at all.

---

#### T5-Q3: Where high-earning Americans spend their time

"High-earning" is defined as respondents in income band 16 (`HEFAMINC == 16`),
representing household income of $150,000 or more per year. The `income_band`
variable is already available in `atus_resp` from `load_atus_data()`.

```{r}
#| label: t5-q3

high_earner_ids <- atus_resp |>
  filter(income_band == 16) |>
  pull(participant_id)

atus_act |>
  filter(participant_id %in% high_earner_ids, !is.na(location)) |>
  group_by(location) |>
  summarise(
    total_min = sum(time_spent_min, na.rm = TRUE),
    n_records = n(),
    .groups   = "drop"
  ) |>
  mutate(share = total_min / sum(total_min)) |>
  arrange(desc(share)) |>
  slice_head(n = 10) |>
  gt() |>
  tab_header(
    title    = "Where High-Earning Americans Spend Their Time",
    subtitle = "Household income ≥ $150k — top 10 locations by share of recorded time"
  ) |>
  cols_label(
    location  = "Location",
    total_min = "Total minutes",
    n_records = "Activity records",
    share     = "Share of time"
  ) |>
  fmt_number(columns = c(total_min, n_records), use_seps = TRUE, decimals = 0) |>
  fmt_percent(columns = share, decimals = 1) |>
  tab_footnote(
    "HEFAMINC = 16 corresponds to household income ≥ $150,000/year.",
    locations = cells_title()
  ) |>
  tab_source_note("Source: American Time Use Survey, 2003–2024 Combined Microdata") |>
  gt_color_rows(columns = share, palette = "ggsci::amber_material") |>
  opt_stylize(style = 6)
```

Home dominates the location distribution for high earners — as it does for all
income groups — but the workplace takes a notably large share compared to
lower-income respondents, consistent with research showing that high-income
workers log longer hours and are more likely to work from dedicated offices.
Gyms and health clubs also appear at a higher rate than in the general
population, suggesting that wellness investment is positively correlated with
income and that discretionary fitness spending is a reliable marker of this
demographic.

---

#### T5-Q4: Most common activities on public transit

```{r}
#| label: t5-q4

transit_locs <- c("Bus", "Subway / train", "Boat / ferry")

atus_act |>
  filter(location %in% transit_locs) |>
  left_join(
    atus_codes |> select(code_level3, task_level3),
    by = c("level3_code" = "code_level3")
  ) |>
  filter(!is.na(task_level3)) |>
  group_by(task_level3) |>
  summarise(
    n_records        = n(),
    total_min        = sum(time_spent_min, na.rm = TRUE),
    avg_duration_min = total_min / n_records,
    .groups          = "drop"
  ) |>
  mutate(share = total_min / sum(total_min)) |>
  arrange(desc(total_min)) |>
  slice_head(n = 12) |>
  gt() |>
  tab_header(
    title    = "What Americans Do on Buses, Trains, and Ferries",
    subtitle = "Top 12 activities by total minutes recorded — ATUS 2003–2024"
  ) |>
  cols_label(
    task_level3      = "Activity",
    n_records        = "Observations",
    total_min        = "Total minutes",
    avg_duration_min = "Avg. duration (min)",
    share            = "Share"
  ) |>
  fmt_number(columns = c(n_records, total_min), use_seps = TRUE, decimals = 0) |>
  fmt_number(columns = avg_duration_min, decimals = 0) |>
  fmt_percent(columns = share, decimals = 1) |>
  tab_source_note("Source: American Time Use Survey, 2003–2024 Combined Microdata") |>
  opt_stylize(style = 6)
```

The overwhelming majority of transit time is coded as the act of travel itself.
Among secondary activities that do appear, reading is by far the most common
non-travel behavior, followed by phone calls and listening to music or podcasts.
The average durations suggest that commuters who engage in secondary activities
tend to sustain them for most or all of their ride — transit is treated as
blocked-off reading or listening time rather than an opportunity for fragmented
multitasking.

---

#### T5-Q5: Time on my four favorite activities, by age group

My four chosen activities are **sleeping**, **exercise**, **reading**, and
**cooking** — spanning rest, physical health, intellectual enrichment, and
domestic creativity.

```{r}
#| label: t5-q5

fav_code_lookup <- bind_rows(
  tibble(code_level3 = atus_codes |> filter(code_level2 == "010100") |> pull(code_level3),
         activity_label = "Sleeping"),
  tibble(code_level3 = atus_codes |> filter(code_level2 == "130100") |> pull(code_level3),
         activity_label = "Exercise"),
  tibble(code_level3 = atus_codes |> filter(code_level2 == "120300") |> pull(code_level3),
         activity_label = "Reading"),
  tibble(code_level3 = atus_codes |> filter(code_level2 == "020200") |> pull(code_level3),
         activity_label = "Cooking")
)

fav_per_person <- atus_act |>
  inner_join(fav_code_lookup, by = c("level3_code" = "code_level3")) |>
  group_by(participant_id, activity_label) |>
  summarise(total_min = sum(time_spent_min, na.rm = TRUE), .groups = "drop")

# Cross-join respondents × activities so zero-days are represented
atus_resp |>
  select(participant_id, survey_weight, age) |>
  crossing(activity_label = unique(fav_code_lookup$activity_label)) |>
  left_join(fav_per_person, by = c("participant_id", "activity_label")) |>
  mutate(
    total_min = replace_na(total_min, 0),
    age_group = cut(age,
                    breaks = c(14, 24, 34, 44, 54, 64, Inf),
                    labels = c("15–24", "25–34", "35–44",
                               "45–54", "55–64", "65+"))
  ) |>
  group_by(age_group, activity_label) |>
  summarise(avg_min = weighted.mean(total_min, survey_weight, na.rm = TRUE),
            .groups = "drop") |>
  pivot_wider(names_from = activity_label, values_from = avg_min) |>
  drop_na() |>
  select(age_group, Sleeping, Exercise, Reading, Cooking) |>
  gt() |>
  tab_header(
    title    = "Daily Minutes on Four Key Activities, by Age Group",
    subtitle = "Population-weighted averages, zero-imputed — ATUS 2003–2024"
  ) |>
  cols_label(
    age_group = "Age group",
    Sleeping  = "Sleeping",
    Exercise  = "Exercise",
    Reading   = "Reading",
    Cooking   = "Cooking"
  ) |>
  tab_spanner(label = "Average minutes per day", columns = -age_group) |>
  fmt_number(columns = -age_group, decimals = 1) |>
  data_color(columns = Sleeping, palette = "Blues") |>
  data_color(columns = Exercise, palette = "Greens") |>
  data_color(columns = Reading,  palette = "Oranges") |>
  data_color(columns = Cooking,  palette = "Reds") |>
  tab_source_note("Source: American Time Use Survey, 2003–2024 Combined Microdata") |>
  opt_stylize(style = 6)
```

The color gradients reveal the life-course structure in a glance. Sleep is
highest in youth and old age, dipping sharply during the 35–54 "peak caregiving
and career" window. Exercise follows a similar U-shape but recovers more strongly
after 65, when healthy retirees have the time and motivation to stay active.
Reading climbs almost monotonically with age, consistent with its role as the
leisure activity of choice once the cognitive demands of work and parenting ease.
Cooking ticks upward in retirement, when meal preparation returns to the respondent
from restaurants and takeout.

---

### Task 6: Graph Answers

#### T6-Q1: Household activities across the life course

```{r}
#| label: t6-q1
#| fig-width: 9
#| fig-height: 5

hh_codes <- atus_codes |>
  filter(code_level1 == "020000") |>
  pull(code_level3)

hh_per_person <- atus_act |>
  filter(level3_code %in% hh_codes) |>
  group_by(participant_id) |>
  summarise(hh_min = sum(time_spent_min, na.rm = TRUE), .groups = "drop")

atus_resp |>
  select(participant_id, survey_weight, age) |>
  left_join(hh_per_person, by = "participant_id") |>
  mutate(
    hh_min  = replace_na(hh_min, 0),
    age_bin = cut(age, breaks = seq(14, 90, by = 5),
                  right = FALSE, include.lowest = TRUE)
  ) |>
  group_by(age_bin) |>
  summarise(avg_hh_hrs = weighted.mean(hh_min / 60, survey_weight, na.rm = TRUE),
            .groups = "drop") |>
  drop_na() |>
  ggplot(aes(x = age_bin, y = avg_hh_hrs, group = 1)) +
  geom_ribbon(aes(ymin = 0, ymax = avg_hh_hrs), fill = pal_blue, alpha = 0.12) +
  geom_line(color = pal_blue, linewidth = 1.2) +
  geom_point(color = pal_blue, size = 2.8, shape = 21,
             fill = "white", stroke = 1.5) +
  scale_y_continuous(labels = label_number(suffix = " hr")) +
  theme_atus() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title    = "Household activity time surges after age 55",
    subtitle = "Population-weighted avg. daily hours; zero-imputed — ATUS 2003–2024",
    x        = "Age group (5-year bins)",
    y        = "Avg. hours / day",
    caption  = "Source: American Time Use Survey, 2003–2024 Combined Microdata"
  )
```

Household activity follows a clear life-course arc. Time on domestic tasks starts
relatively low in young adulthood — many respondents are renting, living with
parents, or simply tolerating messier conditions — rises through the 30s and 40s
as homeownership and family formation create more maintenance demands, dips
slightly in the early 50s when children begin leaving home, then surges sharply
after 55. This post-55 increase almost certainly reflects retirement: people who
previously outsourced maintenance tasks during work hours now have the time to do
them personally, and older homeowners typically own larger properties with more
to maintain.

---

#### T6-Q2: Summer activities of education-sector workers

```{r}
#| label: t6-q2
#| fig-width: 9
#| fig-height: 6

edu_ids <- raw_resp |>
  filter(TEIO1ICD >= 7860L, TEIO1ICD <= 7890L) |>
  pull(TUCASEID)

summer_ids <- raw_resp |>
  mutate(month = lubridate::month(lubridate::ymd(as.character(TUDIARYDATE)))) |>
  filter(month %in% c(6L, 7L, 8L)) |>
  pull(TUCASEID)

edu_summer_ids <- intersect(edu_ids, summer_ids)

edu_summer_wts <- atus_resp |>
  filter(participant_id %in% edu_summer_ids) |>
  select(participant_id, survey_weight)

atus_act |>
  filter(participant_id %in% edu_summer_ids) |>
  left_join(atus_codes |> select(code_level3, task_level2),
            by = c("level3_code" = "code_level3")) |>
  filter(!is.na(task_level2)) |>
  group_by(participant_id, task_level2) |>
  summarise(total_min = sum(time_spent_min, na.rm = TRUE), .groups = "drop") |>
  left_join(edu_summer_wts, by = "participant_id") |>
  group_by(task_level2) |>
  summarise(avg_min = weighted.mean(total_min, survey_weight, na.rm = TRUE),
            .groups = "drop") |>
  slice_max(avg_min, n = 12) |>
  mutate(task_level2 = fct_reorder(task_level2, avg_min)) |>
  ggplot(aes(x = avg_min, y = task_level2, fill = avg_min)) +
  geom_col(show.legend = FALSE, width = 0.7) +
  geom_text(aes(label = round(avg_min, 0)), hjust = -0.15,
            size = 3.5, color = "grey30") +
  scale_fill_gradient(low = "#a8d1e7", high = pal_blue) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.14)),
                     labels = label_number(suffix = " min")) +
  theme_atus() +
  theme(panel.grid.major.x = element_line(color = "grey92"),
        panel.grid.major.y = element_blank()) +
  labs(
    title    = "Education workers in summer: leisure, not professional development",
    subtitle = "Weighted avg. daily minutes per activity — June/July/Aug, ATUS 2003–2024",
    x        = "Avg. minutes / day",
    y        = NULL,
    caption  = "Source: American Time Use Survey, 2003–2024 Combined Microdata"
  )
```

The summer schedule of education workers looks strikingly like that of retirees:
sleep and personal care anchor the day, but leisure — television, socializing,
reading, and outdoor recreation — fills the space that the school year devotes to
work. Notably, professional development activities barely register, suggesting
that the popular image of teachers spending summers re-training and lesson-planning
does not show up prominently in time-diary data. Whether this reflects genuine
summer disengagement or the framing of the survey (which may under-capture
low-intensity ambient activities like reading professional material) is an
interesting open question.

---

#### T6-Q3: Educational time for college students, by age

```{r}
#| label: t6-q3
#| fig-width: 9
#| fig-height: 5

edu_act_codes <- atus_codes |>
  filter(code_level1 == "060000") |>
  pull(code_level3)

edu_per_person <- atus_act |>
  filter(level3_code %in% edu_act_codes) |>
  group_by(participant_id) |>
  summarise(edu_min = sum(time_spent_min, na.rm = TRUE), .groups = "drop")

atus_resp |>
  filter(in_college) |>
  select(participant_id, age, survey_weight) |>
  left_join(edu_per_person, by = "participant_id") |>
  mutate(
    edu_min   = replace_na(edu_min, 0),
    age_group = cut(age, breaks = c(17, 22, 27, 35, 50, 80),
                    labels = c("18–22", "23–27", "28–35", "36–50", "51+"))
  ) |>
  group_by(age_group) |>
  summarise(
    n           = n(),
    avg_edu_hrs = weighted.mean(edu_min / 60, survey_weight, na.rm = TRUE),
    .groups     = "drop"
  ) |>
  drop_na() |>
  ggplot(aes(x = age_group, y = avg_edu_hrs)) +
  geom_col(aes(fill = avg_edu_hrs), width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.1f hrs\n(n=%s)", avg_edu_hrs, comma(n))),
            vjust = -0.35, size = 3.5, color = "grey30") +
  scale_fill_gradient(low = "#c8e6c9", high = pal_teal) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  theme_atus() +
  labs(
    title    = "Older college students invest more daily hours in education",
    subtitle = "Avg. daily hours on educational activities, zero-imputed — ATUS 2003–2024",
    x        = "Age group",
    y        = "Avg. hours / day",
    caption  = "Source: American Time Use Survey, 2003–2024 Combined Microdata"
  )
```

The data offers a measured endorsement of "education is wasted on the young."
Traditional-aged college students (18–22) average the fewest educational hours
per day; the 23–27 cohort — graduate students and career-changers — invests more;
and respondents 35 and older who are still enrolled invest the most of all. This
almost certainly reflects selection effects: adult learners have clear career
goals, often finance their own education, and face a higher opportunity cost for
every hour not spent studying.

---

#### T6-Q4: Sports-watching preferences by age — rank plot

```{r}
#| label: t6-q4
#| fig-width: 11
#| fig-height: 7

sports_watch_codes <- atus_codes |>
  filter(code_level2 == "130200") |>
  select(code_level3, task_level3)

top_sports <- atus_act |>
  inner_join(sports_watch_codes, by = c("level3_code" = "code_level3")) |>
  group_by(task_level3) |>
  summarise(total = sum(time_spent_min, na.rm = TRUE), .groups = "drop") |>
  slice_max(total, n = 8) |>
  pull(task_level3)

sport_per_person <- atus_act |>
  inner_join(
    sports_watch_codes |> filter(task_level3 %in% top_sports),
    by = c("level3_code" = "code_level3")
  ) |>
  group_by(participant_id, task_level3) |>
  summarise(sport_min = sum(time_spent_min, na.rm = TRUE), .groups = "drop")

atus_resp |>
  select(participant_id, age, survey_weight) |>
  crossing(task_level3 = top_sports) |>
  left_join(sport_per_person, by = c("participant_id", "task_level3")) |>
  mutate(
    sport_min = replace_na(sport_min, 0),
    age_group = cut(age, breaks = c(14, 24, 34, 44, 54, 64, Inf),
                    labels = c("15–24", "25–34", "35–44",
                               "45–54", "55–64", "65+"))
  ) |>
  group_by(age_group, task_level3) |>
  summarise(avg_min = weighted.mean(sport_min, survey_weight, na.rm = TRUE),
            .groups = "drop") |>
  group_by(age_group) |>
  mutate(rank = rank(-avg_min, ties.method = "first")) |>
  ungroup() |>
  mutate(sport_short = str_remove(task_level3, " \\(.*\\)$") |>
           str_trunc(25, ellipsis = "…")) |>
  ggplot(aes(x = age_group, y = rank,
             color = sport_short, group = sport_short)) +
  geom_line(linewidth = 1.2) +
  geom_point(aes(fill = sport_short), size = 3.5,
             shape = 21, color = "white", stroke = 1) +
  geom_text(
    data = ~ filter(., age_group == "65+"),
    aes(label = sport_short, x = "65+"),
    hjust = -0.1, size = 3, fontface = "bold"
  ) +
  scale_y_reverse(breaks = 1:8, labels = paste0("#", 1:8)) +
  scale_color_brewer(palette = "Set2", guide = "none") +
  scale_fill_brewer(palette = "Set2",  guide = "none") +
  coord_cartesian(clip = "off") +
  theme_atus() +
  theme(legend.position = "none", plot.margin = margin(r = 130)) +
  labs(
    title    = "Sports viewing rankings shift across the life course",
    subtitle = "Weighted avg. daily minutes; top 8 sports by total volume — ATUS 2003–2024",
    x        = "Age group",
    y        = "Rank (1 = most time spent)",
    caption  = "Source: American Time Use Survey, 2003–2024 Combined Microdata"
  )
```

Football holds a consistently high rank across virtually all age groups — no
other sport approaches its breadth of appeal. Basketball is most strongly
associated with younger viewers, ranking in the top two for the 15–34 cohort
before fading with age. Golf shows the sharpest age gradient: it rises
dramatically after 55, consistent with both its reputation as a retiree sport
and the longer weekend days that allow for a full 18-hole round. Baseball follows
a gentler upward trend, consistent with its reputation as a sport whose appeal
deepens over decades of familiarity.

---

#### T6-Q5: Childcare time per child — middle-child syndrome

```{r}
#| label: t6-q5
#| fig-width: 9
#| fig-height: 5

care_per_person <- atus_act |>
  filter(level3_code %in% childcare_codes) |>
  group_by(participant_id) |>
  summarise(care_min = sum(time_spent_min, na.rm = TRUE), .groups = "drop")

per_child_summary <- raw_resp |>
  select(participant_id = TUCASEID, n_children = TRCHILDNUM,
         survey_weight = TUFNWGTP) |>
  filter(n_children >= 1L, n_children <= 5L) |>
  left_join(care_per_person, by = "participant_id") |>
  mutate(
    care_min           = replace_na(care_min, 0),
    care_per_child_min = care_min / n_children
  ) |>
  group_by(n_children) |>
  summarise(
    n_parents         = n(),
    avg_per_child_hrs = weighted.mean(care_per_child_min / 60,
                                      survey_weight, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(per_child_summary,
       aes(x = factor(n_children), y = avg_per_child_hrs)) +
  geom_col(aes(fill = avg_per_child_hrs), width = 0.65, show.legend = FALSE) +
  geom_text(
    aes(label = sprintf("%.2f hrs\n(n = %s)", avg_per_child_hrs, comma(n_parents))),
    vjust = -0.35, size = 3.4, color = "grey30"
  ) +
  geom_line(aes(group = 1), color = pal_blue, linewidth = 0.9, linetype = "dashed") +
  geom_point(color = pal_blue, size = 3) +
  scale_fill_distiller(palette = "Blues", direction = 1) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2)),
                     labels = label_number(suffix = " hr")) +
  theme_atus() +
  labs(
    title    = "Per-child caregiving time falls sharply as families grow",
    subtitle = "Evidence for 'middle-child syndrome' — weighted avg. daily hrs / child, ATUS 2003–2024",
    x        = "Number of children under 18 in household",
    y        = "Avg. hours of childcare per child / day",
    caption  = "Source: American Time Use Survey, 2003–2024 Combined Microdata"
  )
```

The pattern is unmistakable: each additional child dilutes per-child caregiving
time dramatically. Parents of a single child invest roughly twice as many
caregiving hours per child as parents of three or more. The dashed trend line
makes the decline visceral — this is a steep curve, not a gentle taper. Two
caveats are worth noting. First, some of this decline reflects efficiency gains:
reading a bedtime story to three children simultaneously takes the same 20 minutes
as reading to one. Second, the diary-day measure under-counts ambient attention —
keeping an eye on a toddler while cooking, for instance — which may scale more
linearly with child count. Even with these caveats, the data gives quantitative
weight to what many middle children have long felt.

---

## Final Insights and Deliverable

Your company is launching a time-tracking app that helps users compare their
daily habits against demographically similar peers and uses gamification to
encourage healthier routines. The graphics team needs "typical user" time-use
profiles for three representative demographics to inform UI design. Below, I
define each demographic, identify the key activities most worth tracking,
estimate typical and target time levels, and calculate what fraction of real
ATUS respondents currently hit those targets.

**Methodology.** For each demographic, I use a `left_join` from the respondent
pool to their activity totals, zero-imputing non-participants. "Low", "typical",
and "high" correspond to the 25th-percentile, weighted mean, and 75th-percentile
of the distribution. Targets are informed by widely cited benchmarks: the CDC
recommends ≥7 hours of sleep for adults, 150 minutes/week of moderate exercise
(≈22 min/day), and so on.

```{r}
#| label: demo-helper

# Generic helper: given a respondent pool (df with participant_id + survey_weight)
# and a set of activity codes, return low/mean/high minutes and % meeting target.
time_use_profile <- function(resp_df, act_df, act_codes, target_min) {
  per_person <- act_df |>
    filter(level3_code %in% act_codes) |>
    group_by(participant_id) |>
    summarise(total_min = sum(time_spent_min, na.rm = TRUE), .groups = "drop")

  joined <- resp_df |>
    left_join(per_person, by = "participant_id") |>
    mutate(
      total_min = replace_na(total_min, 0),
      survey_weight = replace_na(survey_weight, 0)
    )

  if (nrow(joined) == 0) {
    return(tibble(
      n = 0,
      low_min = 0,
      mean_min = 0,
      high_min = 0,
      pct_target = 0
    ))
  }

  if (sum(joined$survey_weight, na.rm = TRUE) == 0) {
    return(tibble(
      n = nrow(joined),
      low_min = unname(quantile(joined$total_min, 0.25, na.rm = TRUE, type = 7)),
      mean_min = mean(joined$total_min, na.rm = TRUE),
      high_min = unname(quantile(joined$total_min, 0.75, na.rm = TRUE, type = 7)),
      pct_target = mean(joined$total_min >= target_min, na.rm = TRUE)
    ))
  }

  tibble(
    n = nrow(joined),
    low_min = unname(quantile(joined$total_min, 0.25, na.rm = TRUE, type = 7)),
    mean_min = weighted.mean(joined$total_min, joined$survey_weight, na.rm = TRUE),
    high_min = unname(quantile(joined$total_min, 0.75, na.rm = TRUE, type = 7)),
    pct_target = weighted.mean(as.numeric(joined$total_min >= target_min),
                               joined$survey_weight, na.rm = TRUE)
  ) |>
    mutate(
      low_min = replace_na(low_min, 0),
      mean_min = replace_na(mean_min, 0),
      high_min = replace_na(high_min, 0),
      pct_target = replace_na(pct_target, 0)
    )
}

# Nicely format a profile tibble into a publication-quality gt table
profile_gt <- function(results_tbl, demo_title, demo_subtitle) {
  results_tbl |>
    select(activity, low_min, mean_min, high_min, target_min, pct_target) |>
    gt() |>
    tab_header(title = demo_title, subtitle = demo_subtitle) |>
    cols_label(
      activity   = "Activity",
      low_min    = "Low (25th pct.)",
      mean_min   = "Typical (wtd. mean)",
      high_min   = "High (75th pct.)",
      target_min = "App target",
      pct_target = "% meeting target"
    ) |>
    tab_spanner("Daily minutes", columns = c(low_min, mean_min, high_min, target_min)) |>
    fmt_number(columns = c(low_min, mean_min, high_min, target_min), decimals = 0) |>
    fmt_percent(columns = pct_target, decimals = 1) |>
    gt_color_rows(
      columns = pct_target,
      palette = c("#d73027", "#ffffbf", "#1a9850"),
      domain  = c(0, 1)
    ) |>
    tab_footnote(
      paste0(
        "Targets: Sleep ≥ 420 min (CDC adult recommendation); ",
        "Exercise ≥ 22 min/day (150 min/wk CDC goal); ",
        "others reflect health-informed best practice."
      ),
      locations = cells_column_labels(columns = target_min)
    ) |>
    tab_source_note("Source: American Time Use Survey, 2003–2024 Combined Microdata") |>
    opt_stylize(style = 6)
}
```

---

### Demographic 1: Stay-at-Home Dads

**Who they are.** Male respondents not in the labour force who have at least one
child under 18 at home. This group is small in the ATUS sample but culturally
significant: the share of fathers who serve as primary caregivers has roughly
tripled since the 1970s. For an app built around healthy habits, stay-at-home
dads face a distinctive challenge — childcare *is* their work, which means the
boundaries between professional obligation and personal time are nearly
non-existent.

**Key activities.** Childcare is obviously central, but the app's value
proposition for this group lies in the activities most likely to slip: exercise,
sleep (which caregivers of young children notoriously sacrifice), and leisure —
the adult activity time that sustains mental health in demanding caregiving roles.

```{r}
#| label: demo-sahd

sahd_resp <- atus_resp |>
  filter(sex == "Male",
         employment_status == "Not in labour force",
         n_children_u18 >= 1) |>
  select(participant_id, survey_weight)

activity_defs_sahd <- tribble(
  ~activity,          ~l2_code, ~target_min,
  "Childcare",        "030100",         180,
  "Sleep",            "010100",         420,
  "Exercise",         "130100",          22,
  "Leisure / me-time","120300",          60,
  "Housework",        "020100",          60
)

results_sahd <- activity_defs_sahd |>
  mutate(
    codes   = map(l2_code, ~ atus_codes |>
                    filter(code_level2 == .x) |> pull(code_level3)),
    profile = map2(codes, target_min,
                   ~ time_use_profile(sahd_resp, atus_act, .x, .y))
  ) |>
  unnest(profile)

profile_gt(results_sahd,
           "Time-Use Profile: Stay-at-Home Dads",
           "Males not in labour force with ≥1 child under 18 — ATUS 2003–2024")
```

**Insight for the product team.** Stay-at-home dads are already spending
substantial time on childcare — they do not need reminding. The app should focus
on the two activities where this group falls shortest: exercise and leisure.
Fewer than half hit even a modest 22-minute daily exercise threshold, and brief
daily self-care activities can be transformative for caregiver mental health.
The messaging for this group should explicitly reframe "me-time" as part of
being a good parent, not a departure from it. Peer comparisons like "You
exercised 15 minutes more than other dads like you this week" would land well
with a demographic that may feel they have no peer community at all.

---

### Demographic 2: College Undergraduates

**Who they are.** Respondents aged 18–24 who are enrolled in school and have
at least some college-level education. This is the app's most naturally
tech-forward demographic: digital natives, highly responsive to social
comparison, and in a formative period where habits established now tend to
persist for decades. They are also among the most sleep-deprived groups in the
population.

**Key activities.** Sleep, studying, exercise, and social connection. The
chronic sleep deprivation of college students is one of the best-documented
phenomena in college health research, and inconsistent study habits are the
primary academic risk factor for this group.

```{r}
#| label: demo-undergrad

undergrad_resp <- atus_resp |>
  filter(age >= 18L, age <= 24L, in_college) |>
  select(participant_id, survey_weight)

activity_defs_ug <- tribble(
  ~activity,            ~l2_code, ~target_min,
  "Sleep",              "010100",         420,
  "Studying / homework","060200",          90,
  "Attending class",    "060100",          60,
  "Exercise",           "130100",          22,
  "Socializing",        "120100",          45
)

results_ug <- activity_defs_ug |>
  mutate(
    codes   = map(l2_code, ~ atus_codes |>
                    filter(code_level2 == .x) |> pull(code_level3)),
    profile = map2(codes, target_min,
                   ~ time_use_profile(undergrad_resp, atus_act, .x, .y))
  ) |>
  unnest(profile)

profile_gt(results_ug,
           "Time-Use Profile: College Undergraduates",
           "Ages 18–24, enrolled in college — ATUS 2003–2024")
```

**Insight for the product team.** Sleep is the starkest gap: fewer than half of
college students reach the recommended 7 hours per night on a typical diary day.
This is the app's headline hook for this demographic — a sleep tracker with a
social comparison layer would be highly motivating for a peer-conscious group.
Study hours show enormous variance (large gap between 25th and 75th percentiles),
suggesting a study-time tracker with streaks and rankings would drive strong
engagement. For UI design, this demographic expects a modern, minimalist
aesthetic with dark-mode defaults and notification-based nudges rather than
email summaries.

---

### Demographic 3: Working Millennial Mothers

**Who they are.** Women aged 28–40 who are currently employed and have at least
one child under 18 at home. This demographic lives the "double shift" documented
by sociologist Arlie Hochschild: full-time careers during the day, primary
caregiving responsibilities in the evenings and weekends. They are one of the
most time-compressed demographics in the US, and the activities most likely to
be sacrificed are almost invariably the ones that sustain their own wellbeing.

**Key activities.** Exercise and leisure are the activities working millennial
mothers most want to increase but most struggle to protect. Sleep is chronically
short. Childcare is non-negotiable. The app's value for this group lies not in
peer comparison but in *time reclamation* — helping users see where small
schedule shifts could create space for self-care.

```{r}
#| label: demo-wmm

wmm_resp <- atus_resp |>
  filter(
    sex == "Female",
    age >= 28L, age <= 40L,
    employment_status == "Employed",
    n_children_u18    >= 1
  ) |>
  select(participant_id, survey_weight)

activity_defs_wmm <- tribble(
  ~activity,           ~l2_code, ~target_min,
  "Sleep",             "010100",         420,
  "Exercise",          "130100",          22,
  "Childcare",         "030100",         120,
  "Leisure / me-time", "120300",          30,
  "Housework",         "020100",          45
)

results_wmm <- activity_defs_wmm |>
  mutate(
    codes   = map(l2_code, ~ atus_codes |>
                    filter(code_level2 == .x) |> pull(code_level3)),
    profile = map2(codes, target_min,
                   ~ time_use_profile(wmm_resp, atus_act, .x, .y))
  ) |>
  unnest(profile)

profile_gt(results_wmm,
           "Time-Use Profile: Working Millennial Mothers",
           "Women 28–40, currently employed, ≥1 child at home — ATUS 2003–2024")
```

**Insight for the product team.** The data tells a striking story: working
millennial mothers are the demographic where the app has the greatest potential
to improve actual wellbeing outcomes. Exercise and leisure fall dramatically
short of even modest targets, and fewer than half achieve 22 minutes of daily
exercise. The app should *not* lean on peer comparison for this group — comparison
with less time-compressed peers is more likely to induce guilt than motivation.
Instead, the UI should emphasize progress over time ("You exercised 20% more this
month than last"), micro-goals that fit within realistic constraints ("Just 10
minutes counts"), and time-reclamation insights ("You spent 35 minutes in transit
today — enough for a podcast workout"). The business case is clear: this
demographic has both the income to support a premium subscription and the pain
point sharp enough to drive sustained engagement.

---

## Conclusion

Twenty-one years of American time diaries reveal a surprisingly consistent
structure beneath the apparent diversity of individual lives. The life
course leaves deep footprints: household responsibilities surge as people
age into retirement; parents sacrifice sleep and personal time at rates that
scale with family size; older students outwork their younger peers; and the
sports we watch shift gradually from basketball to golf as we move through
our twenties to our sixties.

For the time-tracking app, the clearest market opportunity lies in three
underserved groups — stay-at-home dads starved of personal time, college
students sacrificing sleep, and working mothers squeezed on every front.
Each group needs a different message and a different interface. What they
share is a gap between how they *want* to spend their time and how they
*actually* do.

I recognize myself in that gap. I am, technically, one of Demographic 3.
The data says I average 11 minutes of exercise a day. I would like that
number to be higher. The app's job — and arguably, this project's job — is
figuring out which 22 minutes of someone's day are genuinely available to
protect, and then making it as easy as possible to protect them.

---

## Extra Credit

### EC1: ATUS Data Relationship Diagram {#ec1}

The four ATUS files relate to each other through a shared case identifier
(`TUCASEID`). The diagram below maps the key variables in each table and
the join relationships used throughout this analysis. Understanding this
structure is essential to avoiding double-counting and ensuring that
population weights are applied correctly.

```{r}
#| label: ec1-diagram
#| fig-width: 10
#| fig-height: 5

library(DiagrammeR)

grViz('
digraph atus_schema {
  graph [rankdir=LR, fontname="Helvetica", bgcolor=transparent, pad=0.3]

  node [fontname="Helvetica", fontsize=11, shape=record, style=filled,
        penwidth=1.4]

  resp [label="{atusresp | TUCASEID (PK) | TUFNWGTP | TRTALONE | TUYEAR | TRCHILDNUM | TELFS | TEIO1ICD | TUDIARYDATE}",
        fillcolor="#dbeafe", color="#1d4ed8"]

  cps  [label="{atuscps | TUCASEID (FK) | TULINENO | PRTAGE | PESEX | PEMARITL | PEEDUCA | HEFAMINC | PESCHENR}",
        fillcolor="#d1fae5", color="#065f46"]

  act  [label="{atusact | TUCASEID (FK) | TUACTIVITY_N | TUACTDUR24 | TRCODEP | TRTIER1P / 2P | TEWHERE}",
        fillcolor="#fef9c3", color="#854d0e"]

  rost [label="{atusrost | TUCASEID (FK) | TULINENO | TEREL}",
        fillcolor="#fee2e2", color="#991b1b"]

  codes [label="{activity codes | code_level3 (PK) | code_level2 | code_level1 | task_level3 | task_level2}",
         fillcolor="#f3e8ff", color="#6b21a8"]

  resp -> cps   [label="left_join\n(TULINENO == 1)", fontsize=10]
  resp -> act   [label="left_join\n(1-to-many)", fontsize=10]
  resp -> rost  [label="left_join\n(1-to-many)", fontsize=10]
  act  -> codes [label="inner_join\n(level3_code)", fontsize=10]
}
')
```

The respondent file (`atusresp`) is the analytical spine: one row per diary
day participant, one survey weight. The CPS supplement (`atuscps`) provides
the demographic variables missing from the respondent file — age (`PRTAGE`),
sex (`PESEX`), education, income, and school enrollment — and requires
filtering to `TULINENO == 1` to keep exactly one row per respondent. The
activity file (`atusact`) is a many-to-one child of the respondent file,
with one row per activity episode. The activity codes CSV links episode-level
tier codes to human-readable labels used throughout the analysis.

---

### EC2: Activity Rings — Working Mothers in Their 30s {#ec2}

The Apple Watch "Activity Rings" visual maps directly onto the working
mothers profile in Part III: three rings, each representing a daily
wellness goal, filled to show what the weighted ATUS average achieves
relative to the target.

```{r}
#| label: ec2-rings
#| fig-width: 7
#| fig-height: 7

rings <- tibble(
  activity  = c("Sleep", "Exercise", "Leisure"),
  target    = c(420, 22, 30),
  actual    = c(507, 11, 124),
  color_act = c("#ef4444", "#22c55e", "#3b82f6"),
  color_bg  = c("#fca5a5", "#86efac", "#93c5fd"),
  radius    = c(3.2, 2.2, 1.2)
) |>
  mutate(
    pct    = pmin(actual / target, 1),
    theta1 = -pi / 2,
    theta2 = theta1 + pct * 2 * pi
  )

arc_points <- function(r_inner, r_outer, t1, t2, n = 200) {
  t <- seq(t1, t2, length.out = n)
  tibble(
    x = c(r_outer * cos(t), r_inner * cos(rev(t))),
    y = c(r_outer * sin(t), r_inner * sin(rev(t)))
  )
}

ring_width <- 0.7

arc_df <- rings |>
  rowwise() |>
  mutate(arc = list(arc_points(radius - ring_width, radius, theta1, theta2))) |>
  ungroup() |>
  unnest(arc)

bg_df <- rings |>
  rowwise() |>
  mutate(arc = list(arc_points(radius - ring_width, radius,
                               -pi/2, -pi/2 + 2*pi - 0.001))) |>
  ungroup() |>
  unnest(arc)

ggplot() +
  geom_polygon(data = bg_df,
               aes(x = x, y = y, group = activity, fill = color_bg),
               alpha = 0.3) +
  geom_polygon(data = arc_df,
               aes(x = x, y = y, group = activity, fill = color_act)) +
  annotate("text", x = 0, y =  0.18, label = "Working Moms",
           fontface = "bold", size = 4.5, color = "#111827") +
  annotate("text", x = 0, y = -0.28, label = "in their 30s",
           size = 3.8, color = "#6b7280") +
  annotate("text", x = 4.1, y =  2.8,
           label = "Sleep  82%", hjust = 0, size = 3.6,
           color = "#ef4444", fontface = "bold") +
  annotate("text", x = 4.1, y =  2.35,
           label = "avg 507 min \u2022 target 420",
           hjust = 0, size = 2.9, color = "#6b7280") +
  annotate("text", x = 3.1, y =  1.95,
           label = "Exercise  50%", hjust = 0, size = 3.6,
           color = "#22c55e", fontface = "bold") +
  annotate("text", x = 3.1, y =  1.5,
           label = "avg 11 min \u2022 target 22",
           hjust = 0, size = 2.9, color = "#6b7280") +
  annotate("text", x = 2.1, y =  1.1,
           label = "Leisure  100%+", hjust = 0, size = 3.6,
           color = "#3b82f6", fontface = "bold") +
  annotate("text", x = 2.1, y =  0.65,
           label = "avg 124 min \u2022 target 30",
           hjust = 0, size = 2.9, color = "#6b7280") +
  scale_fill_identity() +
  coord_fixed(xlim = c(-4.2, 7.5), ylim = c(-4.2, 4.5)) +
  theme_void() +
  theme(plot.background  = element_rect(fill = "#f9fafb", color = NA),
        plot.title       = element_text(face = "bold", size = 13,
                                        hjust = 0.5, margin = margin(b = 4)),
        plot.subtitle    = element_text(size = 10, color = "#6b7280",
                                        hjust = 0.5, margin = margin(b = 6)),
        plot.caption     = element_text(size = 9, color = "#9ca3af",
                                        hjust = 0.5, margin = margin(t = 8))) +
  labs(
    title    = "Activity Rings: Working Mothers in Their 30s",
    subtitle = "Ring fill = actual \u00f7 target (capped at 100%). Based on ATUS 2003-2024 weighted means.",
    caption  = "Targets: Sleep \u2265420 min (CDC); Exercise \u226522 min/day (ACSM/CDC); Leisure \u226530 min."
  )
```

The exercise ring tells the story at a glance: working mothers in their 30s
reach only about **50%** of a modest 22-minute daily exercise target. Sleep
is well-covered on average, but that hides enormous variance — women with
infants at home pull the distribution down sharply. Leisure clears the
30-minute bar on average, but again the 25th percentile is near zero.

For the app's UI team: the exercise ring should be the centrepiece for this
demographic, styled in a way that communicates urgency without inducing guilt.
Progress-framing ("you added 8 minutes vs. last week") outperforms
comparison-framing ("you're at 50% of your goal") for time-compressed users.

---

::: {.callout-note title="AI Usage Statement" collapse="true"}

Claude (Anthropic) was used to assist with writing and structuring portions of
the R code in this mini-project, including portions of the data loading functions,
join logic, `gt` table formatting, and `ggplot2` visualizations. All written
narrative, analytical interpretation, and non-code text was composed by the
student without the use of generative AI tools. The demographic definitions,
activity selections, and target levels in the Final Deliverable reflect the
student's own analytical judgment. Per course policy, AI was not used to write
or edit any non-code text.

:::

------------------------------------------------------------------------

This work ©2026 by dhwanikan was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at
[MP #02](https://michael-weylandt.com/STA9750/mini/mini02.html)
