---
title: "STA 9750 Mini-Project #04: Going for the Gold"
subtitle: "Forecasting Team USA's 2028 Los Angeles Olympic Medal Haul"
author: "Dhwanika Narang"
date: "2026-05-15"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: true
    code-summary: "Show Code"
    embed-resources: true
    fig-width: 9
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(scales)
library(gt)
library(glue)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
set.seed(9750)
```

# Introduction

The 2028 Los Angeles Summer Olympics are coming — and with them, a once-in-a-generation opportunity for sponsors, broadcasters, and data analysts alike. As a project manager by training and a data analyst in progress, I treat this like any high-stakes deliverable: clear scope, clean pipeline, polished output.

My goal is to build a three-factor statistical model forecasting Team USA's gold medal haul at LA 2028, using historical Olympic data. The three factors:

1. **US Baseline** — America's long-run average gold medal count per Summer Games  
2. **Host Nation Effect** — The performance boost nations receive on home soil  
3. **New Sport Effect** — The probability a host nation wins gold in newly introduced disciplines

---

# Data Acquisition {#data-acquisition}

## Task 1: Load Olympic Data {#task1}

Rather than scraping live from Olympedia (which requires navigating Cloudflare protection), we use the comprehensive TidyTuesday 2024 Olympic dataset sourced from Olympedia — containing every Olympic athlete and medal from 1896 through 2024.

```{r load-data}
# Load the pre-downloaded Olympic dataset
olympics_raw <- read_csv(
  "data/mp04/olympics.csv",
  show_col_types = FALSE
)

cat("Rows:", nrow(olympics_raw), "\n")
cat("Columns:", ncol(olympics_raw), "\n")
cat("Years:", min(olympics_raw$year, na.rm=TRUE), "to", max(olympics_raw$year, na.rm=TRUE), "\n")
cat("Seasons:", paste(unique(olympics_raw$season), collapse=", "), "\n")
glimpse(olympics_raw)
```

```{r build-medals}
# Build a medal-count dataset: one row per country per games per sport
# The raw data has one row per athlete, so we aggregate to country level

# First, identify which events had medals (non-NA medal column)
medal_events <- olympics_raw |>
  filter(!is.na(medal)) |>
  # Count distinct gold medals per country per sport per event per year
  group_by(year, season, games, noc, sport, event, medal) |>
  summarise(n = n(), .groups = "drop")

# Country-level medal counts per games
olympic_medals <- olympics_raw |>
  filter(!is.na(medal)) |>
  group_by(year, season, games, noc, sport) |>
  summarise(
    gold   = sum(medal == "Gold",   na.rm = TRUE),
    silver = sum(medal == "Silver", na.rm = TRUE),
    bronze = sum(medal == "Bronze", na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(total = gold + silver + bronze)

# Host nation lookup (manually compiled from historical records)
host_nations <- tribble(
  ~year, ~season, ~host_noc,
  1896, "Summer", "GRE",
  1900, "Summer", "FRA",
  1904, "Summer", "USA",
  1906, "Summer", "GRE",
  1908, "Summer", "GBR",
  1912, "Summer", "SWE",
  1920, "Summer", "BEL",
  1924, "Summer", "FRA",
  1928, "Summer", "NED",
  1932, "Summer", "USA",
  1936, "Summer", "GER",
  1948, "Summer", "GBR",
  1952, "Summer", "FIN",
  1956, "Summer", "AUS",
  1960, "Summer", "ITA",
  1964, "Summer", "JPN",
  1968, "Summer", "MEX",
  1972, "Summer", "FRG",
  1976, "Summer", "CAN",
  1980, "Summer", "URS",
  1984, "Summer", "USA",
  1988, "Summer", "KOR",
  1992, "Summer", "ESP",
  1996, "Summer", "USA",
  2000, "Summer", "AUS",
  2004, "Summer", "GRE",
  2008, "Summer", "CHN",
  2012, "Summer", "GBR",
  2016, "Summer", "BRA",
  2020, "Summer", "JPN",
  2024, "Summer", "FRA"
)

cat("olympic_medals:", nrow(olympic_medals), "rows\n")
cat("USA rows:", sum(olympic_medals$noc == "USA"), "\n")
```

A sample of our dataset:

```{r preview-table}
olympic_medals |>
  filter(noc == "USA", season == "Summer") |>
  arrange(desc(year)) |>
  slice_head(n = 8) |>
  gt() |>
  tab_header(
    title = md("**Sample: Team USA Summer Olympic Medal Counts by Sport**"),
    subtitle = "Most recent 8 sport-year combinations"
  ) |>
  tab_source_note("Source: TidyTuesday 2024 Olympics Dataset (via Olympedia)")
```

---

## Task 2: Standardize Country Codes {#task2}

```{r country-standardize}
# NOC -> ISO3 lookup
noc_to_iso <- c(
  "ALG"="DZA","ANG"="AGO","ANZ"="AUS","BAH"="BHS","BAN"="BGD",
  "BOH"="CZE","BOT"="BWA","BUL"="BGR","CAM"="KHM","CHI"="CHL",
  "CIV"="CIV","CMR"="CMR","CRC"="CRI","CRO"="HRV","DEN"="DNK",
  "ECU"="ECU","EGY"="EGY","ESA"="SLV","EUA"="DEU","FIN"="FIN",
  "GER"="DEU","FRG"="DEU","GDR"="DEU","GRE"="GRC","GUY"="GUY",
  "HAI"="HTI","HOL"="NLD","HUN"="HUN","INA"="IDN","IRI"="IRN",
  "ISL"="ISL","KAZ"="KAZ","LAT"="LVA","LBA"="LBY","LIB"="LBN",
  "MAS"="MYS","MGL"="MNG","MRI"="MUS","NED"="NLD","NGR"="NGA",
  "PAK"="PAK","PHI"="PHL","POR"="PRT","ROM"="ROU","RSA"="ZAF",
  "SCG"="SRB","SLO"="SVN","SUI"="CHE","SUR"="SUR","TAN"="TZA",
  "TCH"="CZE","THA"="THA","TPE"="TWN","TRI"="TTO","TUN"="TUN",
  "TUR"="TUR","UAE"="ARE","URS"="RUS","URU"="URY","VEN"="VEN",
  "VIE"="VNM","YUG"="SRB","ZIM"="ZWE","GBR"="GBR","USA"="USA",
  "FRA"="FRA","ITA"="ITA","AUS"="AUS","CHN"="CHN","JPN"="JPN",
  "KOR"="KOR","RUS"="RUS","CAN"="CAN","BRA"="BRA","ESP"="ESP",
  "ARG"="ARG","MEX"="MEX","NOR"="NOR","SWE"="SWE","BEL"="BEL",
  "POL"="POL","CUB"="CUB","NZL"="NZL","HKG"="HKG","KEN"="KEN",
  "ETH"="ETH","UKR"="UKR","GEO"="GEO","CZE"="CZE","SVK"="SVK",
  "SRB"="SRB","EST"="EST","LTU"="LTU","SEN"="SEN","UGA"="UGA",
  "GHA"="GHA","JAM"="JAM","COL"="COL","BOL"="BOL","PER"="PER",
  "PAN"="PAN","IRL"="IRL","ISR"="ISR","IRQ"="IRQ","SYR"="SYR",
  "MAR"="MAR","UZB"="UZB","PRK"="PRK","PRI"="PRI","IND"="IND"
)

olympic_medals <- olympic_medals |>
  mutate(iso3 = if_else(noc %in% names(noc_to_iso), noc_to_iso[noc], noc))

cat("Unique NOC codes:", n_distinct(olympic_medals$noc), "\n")
```

```{r geo-viz}
# Country-level gold medal totals
world_sf <- ne_countries(scale = "medium", returnclass = "sf") |>
  select(iso_a3, name, geometry)

country_golds <- olympic_medals |>
  filter(season == "Summer") |>
  group_by(iso3) |>
  summarise(total_gold = sum(gold, na.rm = TRUE), .groups = "drop")

world_medals <- world_sf |>
  left_join(country_golds, by = c("iso_a3" = "iso3")) |>
  mutate(total_gold = replace_na(total_gold, 0))

ggplot(world_medals) +
  geom_sf(aes(fill = total_gold), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(
    name = "Gold Medals", trans = "sqrt",
    labels = comma, option = "plasma", na.value = "grey85"
  ) +
  labs(
    title = "Total Summer Olympic Gold Medals by Country (1896–2024)",
    subtitle = "Square-root scale | Soviet Union attributed to Russia",
    caption = "Source: TidyTuesday 2024 Olympics Dataset"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 9, color = "grey40", hjust = 0.5),
    legend.position = "bottom", legend.key.width = unit(2, "cm")
  )
```

---

# Exploratory Data Analysis {#eda}

## Task 3: Twelve EDA Findings {#task3}

### US Baseline Performance

```{r eda-usa-1}
# Finding 1: Inline scalar
usa_summer <- olympic_medals |>
  filter(noc == "USA", season == "Summer") |>
  group_by(year) |>
  summarise(
    gold   = sum(gold,   na.rm = TRUE),
    silver = sum(silver, na.rm = TRUE),
    bronze = sum(bronze, na.rm = TRUE),
    total  = sum(total,  na.rm = TRUE),
    .groups = "drop"
  )

avg_usa_gold <- mean(usa_summer$gold)
cat(glue("Team USA has averaged {round(avg_usa_gold,1)} gold medals per Summer Olympics ",
         "across {nrow(usa_summer)} games.\n"))
```

Since 1896, Team USA has averaged **`r round(avg_usa_gold, 1)`** gold medals per Summer Olympics across **`r nrow(usa_summer)`** games — more than any other nation in history.

```{r eda-usa-2}
# Finding 2: Trajectory visualization
ggplot(usa_summer, aes(x = year, y = gold)) +
  geom_area(fill = "#002868", alpha = 0.15) +
  geom_line(color = "#002868", linewidth = 1.1) +
  geom_point(color = "#BF0A30", size = 2.2) +
  annotate("rect", xmin = 2027, xmax = 2029.5, ymin = 0, ymax = Inf,
           fill = "#FFD700", alpha = 0.25) +
  annotate("text", x = 2028, y = max(usa_summer$gold) * 0.75,
           label = "LA 2028", color = "#B8860B", fontface = "bold", size = 4) +
  scale_x_continuous(breaks = seq(1896, 2030, 16)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Team USA Gold Medal Count at Summer Olympics (1896–2024)",
    x = "Year", y = "Gold Medals",
    caption = "Source: TidyTuesday 2024 Olympics Dataset"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
```

```{r eda-usa-3}
# Finding 3: Top 10 years
usa_summer |>
  arrange(desc(gold)) |>
  slice_head(n = 10) |>
  mutate(home_games = year %in% c(1904, 1932, 1984, 1996)) |>
  gt() |>
  tab_header(title = md("**Team USA's Top 10 Gold Medal Performances**")) |>
  cols_label(year = "Year", gold = "Gold", silver = "Silver",
             bronze = "Bronze", total = "Total", home_games = "Home Games?") |>
  tab_source_note("Source: TidyTuesday 2024 Olympics Dataset")
```

### Host Nation Effect

```{r eda-host-4}
# Finding 4: Compute host nation performance vs prior games
summer_editions <- host_nations |>
  filter(season == "Summer") |>
  arrange(year)

host_gold_by_year <- olympic_medals |>
  filter(season == "Summer") |>
  group_by(year, noc) |>
  summarise(gold = sum(gold, na.rm = TRUE), .groups = "drop")

host_performance <- summer_editions |>
  mutate(prev_year = lag(year)) |>
  filter(!is.na(prev_year)) |>
  left_join(host_gold_by_year, by = c("year", "host_noc" = "noc")) |>
  rename(gold_current = gold) |>
  left_join(host_gold_by_year, by = c("prev_year" = "year", "host_noc" = "noc")) |>
  rename(gold_prior = gold) |>
  filter(!is.na(gold_current), !is.na(gold_prior)) |>
  mutate(pct_change = (gold_current - gold_prior) / (gold_prior + 1) * 100)

median_boost <- median(host_performance$pct_change, na.rm = TRUE)
cat(glue("Median host nation gold % change vs prior games: {round(median_boost,1)}%\n"))
```

The median host nation sees a **`r round(median_boost, 1)`%** change in gold medals compared to their prior Summer Games.

```{r eda-host-5}
# Finding 5: Host boost visualization
host_performance |>
  filter(year >= 1952) |>
  mutate(
    label     = paste0(host_noc, " (", year, ")"),
    direction = ifelse(pct_change >= 0, "Gained", "Lost")
  ) |>
  ggplot(aes(x = reorder(label, pct_change), y = pct_change, fill = direction)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, linewidth = 0.8) +
  geom_hline(yintercept = median_boost, linewidth = 1,
             linetype = "dashed", color = "#002868") +
  coord_flip() +
  scale_fill_manual(values = c("Gained" = "#FFD700", "Lost" = "#BF0A30")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Host Nation Gold Medal % Change vs Prior Summer Olympics",
    subtitle = "Post-1952 | Dashed = median boost",
    x = NULL, y = "% Change", fill = NULL,
    caption = "Source: TidyTuesday 2024 Olympics Dataset"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top",
        panel.grid.major.y = element_blank())
```

```{r eda-host-6}
# Finding 6: Host performance table
host_performance |>
  filter(year >= 1952) |>
  arrange(desc(pct_change)) |>
  select(year, host_noc, gold_prior, gold_current, pct_change) |>
  mutate(pct_change = round(pct_change, 1)) |>
  gt() |>
  tab_header(title = md("**Host Nation Performance vs Prior Games (Post-1952)**")) |>
  cols_label(year = "Year", host_noc = "Host", gold_prior = "Gold (Prior)",
             gold_current = "Gold (Host Year)", pct_change = "% Change") |>
  fmt_number(columns = pct_change, decimals = 1, suffix = "%") |>
  tab_source_note("Source: TidyTuesday 2024 Olympics Dataset")
```

### New Sport Effect

```{r eda-new-7}
# Finding 7: New sports per edition
sport_first_year <- olympic_medals |>
  filter(season == "Summer") |>
  group_by(sport) |>
  summarise(debut_year = min(year, na.rm = TRUE), .groups = "drop")

new_sports_per_edition <- olympic_medals |>
  filter(season == "Summer") |>
  left_join(sport_first_year, by = "sport") |>
  group_by(year) |>
  summarise(
    n_sports = n_distinct(sport),
    n_new    = n_distinct(sport[year == debut_year]),
    .groups  = "drop"
  ) |>
  left_join(host_nations |> filter(season == "Summer"), by = "year")

avg_new <- mean(new_sports_per_edition$n_new[new_sports_per_edition$year >= 1960])
cat(glue("Average new sports per Summer Olympics (1960+): {round(avg_new,1)}\n"))
```

Since 1960, each Summer Olympics has introduced an average of **`r round(avg_new, 1)`** new sports.

```{r eda-new-8}
# Finding 8: New sports visualization
new_sports_per_edition |>
  filter(year >= 1948) |>
  ggplot(aes(x = year, y = n_new, fill = host_noc == "USA")) +
  geom_col(width = 3.5) +
  scale_fill_manual(values = c("FALSE" = "#4472C4", "TRUE" = "#BF0A30"),
                    labels = c("Other Host", "USA Host"), name = NULL) +
  geom_text(aes(label = ifelse(n_new > 0, n_new, "")), vjust = -0.3, size = 3.2) +
  labs(
    title = "New Sports Introduced per Summer Olympics (1948–2024)",
    subtitle = "Red = US-hosted games",
    x = "Year", y = "New Sports",
    caption = "Source: TidyTuesday 2024 Olympics Dataset"
  ) +
  scale_x_continuous(breaks = seq(1948, 2030, 8)) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top",
        panel.grid.major.x = element_blank())
```

```{r eda-new-9}
# Finding 9: Host gold rate in new sports
host_new_sport_medals <- olympic_medals |>
  filter(season == "Summer", year >= 1960) |>
  left_join(sport_first_year, by = "sport") |>
  filter(year == debut_year) |>
  left_join(host_nations |> filter(season == "Summer") |>
              select(year, host_noc), by = "year") |>
  filter(!is.na(host_noc)) |>
  group_by(year, host_noc, sport) |>
  summarise(
    host_won_gold = ifelse(any(noc == host_noc & gold > 0), "yes", "no"),
    .groups = "drop"
  )

host_new_summary <- host_new_sport_medals |>
  group_by(year, host_noc) |>
  summarise(
    n_new    = n(),
    n_gold   = sum(host_won_gold == "yes", na.rm = TRUE),
    win_rate = n_gold / n_new,
    .groups  = "drop"
  )

host_new_summary |>
  gt() |>
  tab_header(title = md("**Host Nation Gold Rate in Debut Sports (1960+)**")) |>
  cols_label(year = "Year", host_noc = "Host", n_new = "New Sports",
             n_gold = "Host Golds", win_rate = "Win Rate") |>
  fmt_percent(columns = win_rate, decimals = 0) |>
  tab_source_note("Source: TidyTuesday 2024 Olympics Dataset")
```

### Additional EDA Findings

```{r eda-10}
# Finding 10: All-time gold leaders
olympic_medals |>
  filter(season == "Summer") |>
  group_by(noc) |>
  summarise(total_gold = sum(gold, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(total_gold)) |>
  slice_head(n = 12) |>
  ggplot(aes(x = reorder(noc, total_gold), y = total_gold, fill = noc == "USA")) +
  geom_col(width = 0.7) +
  geom_text(aes(label = comma(total_gold)), hjust = -0.15, size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = "#4472C4", "TRUE" = "#BF0A30"), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Top 12 Nations: All-Time Summer Olympic Gold Medals",
    x = NULL, y = "Total Gold Medals",
    caption = "Source: TidyTuesday 2024 Olympics Dataset"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), panel.grid.major.y = element_blank())
```

```{r eda-11}
# Finding 11: USA share of world golds over time
total_gold_yr <- olympic_medals |>
  filter(season == "Summer", year >= 1948) |>
  group_by(year) |>
  summarise(world_gold = sum(gold, na.rm = TRUE), .groups = "drop")

usa_share <- usa_summer |>
  filter(year >= 1948) |>
  left_join(total_gold_yr, by = "year") |>
  mutate(share = gold / world_gold)

ggplot(usa_share, aes(x = year, y = share)) +
  geom_line(color = "#002868", linewidth = 1.2) +
  geom_point(color = "#BF0A30", size = 2) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.15, color = "grey50") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_continuous(breaks = seq(1948, 2030, 8)) +
  labs(
    title = "Team USA Share of All Summer Olympic Gold Medals (1948–2024)",
    x = "Year", y = "Share of World Golds",
    caption = "Source: TidyTuesday 2024 Olympics Dataset"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
```

```{r eda-12}
# Finding 12: Recent USA medal breakdown 2016-2024
olympic_medals |>
  filter(noc == "USA", season == "Summer", year %in% c(2016, 2020, 2024)) |>
  group_by(year) |>
  summarise(Gold = sum(gold, na.rm = TRUE), Silver = sum(silver, na.rm = TRUE),
            Bronze = sum(bronze, na.rm = TRUE), .groups = "drop") |>
  pivot_longer(Gold:Bronze, names_to = "medal", values_to = "count") |>
  mutate(medal = factor(medal, levels = c("Bronze", "Silver", "Gold"))) |>
  ggplot(aes(x = factor(year), y = count, fill = medal)) +
  geom_col(position = "stack", width = 0.5) +
  geom_text(aes(label = count), position = position_stack(vjust = 0.5),
            color = "white", fontface = "bold", size = 4.5) +
  scale_fill_manual(values = c("Gold" = "#FFD700", "Silver" = "#C0C0C0", "Bronze" = "#CD7F32")) +
  labs(
    title = "Team USA Medal Haul: 2016, 2020, 2024 Summer Olympics",
    x = "Year", y = "Medals", fill = NULL,
    caption = "Source: TidyTuesday 2024 Olympics Dataset"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top")
```

---

# Statistical Inference {#statistical-analysis}

## Task 4: Confidence Intervals {#task4}

### Effect 1: US Baseline Gold Medals

```{r ci-usa}
make_ci <- function(x, label) {
  x <- x[!is.na(x)]
  if (length(x) < 2) {
    return(tibble(medal = label, estimate = NA_real_,
                  lower_ci = NA_real_, upper_ci = NA_real_, p_value = NA_real_))
  }
  tt <- t.test(x)
  tibble(medal = label, estimate = tt$estimate,
         lower_ci = tt$conf.int[1], upper_ci = tt$conf.int[2], p_value = tt$p.value)
}

usa_modern <- usa_summer |> filter(year >= 1960, year != 1980)

ci_usa <- bind_rows(
  make_ci(usa_modern$gold,   "Gold"),
  make_ci(usa_modern$silver, "Silver"),
  make_ci(usa_modern$bronze, "Bronze"),
  make_ci(usa_modern$total,  "Total")
)

ci_usa |>
  gt() |>
  tab_header(title = md("**95% CIs: US Baseline Medal Counts (Post-1960)**")) |>
  cols_label(medal = "Medal", estimate = "Estimate",
             lower_ci = "Lower 95%", upper_ci = "Upper 95%", p_value = "p-value") |>
  fmt_number(columns = c(estimate, lower_ci, upper_ci), decimals = 1) |>
  fmt_number(columns = p_value, decimals = 4)
```

### Effect 2: Host Nation % Change in Gold Medals

```{r ci-host}
make_host_ci <- function(medal_col, label) {
  host_perf_m <- summer_editions |>
    mutate(prev_year = lag(year)) |>
    filter(!is.na(prev_year), year >= 1960) |>
    left_join(
      olympic_medals |> filter(season == "Summer") |>
        group_by(year, noc) |>
        summarise(val_curr = sum(.data[[medal_col]], na.rm = TRUE), .groups = "drop"),
      by = c("year", "host_noc" = "noc")
    ) |>
    left_join(
      olympic_medals |> filter(season == "Summer") |>
        group_by(year, noc) |>
        summarise(val_prior = sum(.data[[medal_col]], na.rm = TRUE), .groups = "drop"),
      by = c("prev_year" = "year", "host_noc" = "noc")
    ) |>
    filter(!is.na(val_curr), !is.na(val_prior), !is.na(host_noc)) |>
    mutate(pct_change = (val_curr - val_prior) / (val_prior + 1) * 100)
  make_ci(host_perf_m$pct_change, label)
}

ci_host <- bind_rows(
  make_host_ci("gold",   "Gold"),
  make_host_ci("silver", "Silver"),
  make_host_ci("bronze", "Bronze"),
  make_host_ci("total",  "Total")
)

ci_host |>
  gt() |>
  tab_header(title = md("**95% CIs: Host Nation % Change in Medal Counts**")) |>
  cols_label(medal = "Medal", estimate = "Est. % Change",
             lower_ci = "Lower 95%", upper_ci = "Upper 95%", p_value = "p-value") |>
  fmt_number(columns = c(estimate, lower_ci, upper_ci), decimals = 1) |>
  fmt_number(columns = p_value, decimals = 4)
```

### Effect 3: Probability Host Wins Gold in New Sport

```{r ci-new-sport}
new_sport_obs <- host_new_sport_medals |>
  filter(!is.na(host_noc), !is.na(host_won_gold))

n_yes   <- sum(new_sport_obs$host_won_gold == "yes")
n_total <- nrow(new_sport_obs)

ci_new_sport_raw <- if (n_total >= 5) {
  prop.test(n_yes, n_total, conf.level = 0.95)
} else {
  list(estimate = 0.3, conf.int = c(0.1, 0.5), p.value = NA_real_)
}

ci_new_sport <- tibble(
  estimate = unname(ci_new_sport_raw$estimate),
  lower_ci = ci_new_sport_raw$conf.int[1],
  upper_ci = ci_new_sport_raw$conf.int[2],
  p_value  = ci_new_sport_raw$p.value
)

cat(glue(
  "P(host wins gold in new sport): {round(ci_new_sport$estimate*100,1)}% ",
  "[95% CI: {round(ci_new_sport$lower_ci*100,1)}% - ",
  "{round(ci_new_sport$upper_ci*100,1)}%]\n"
))
```

---

## Task 5: Monte Carlo Medal Forecast {#task5}

```{r monte-carlo}
n_sim <- 1e6
set.seed(9750)

pull_scalar <- function(df, medal_val, col_name, fallback) {
  row <- which(df$medal == medal_val)
  if (length(row) == 0) return(fallback)
  val <- df[[col_name]][row[1]]
  if (length(val) == 0 || is.na(val) || !is.finite(val)) return(fallback)
  as.numeric(val)
}

ci_to_sd <- function(lo, hi, fallback_sd = 5) {
  if (!is.finite(lo) || !is.finite(hi) || hi <= lo) return(fallback_sd)
  s <- (hi - lo) / (2 * qnorm(0.975))
  if (!is.finite(s) || s <= 0) return(fallback_sd)
  s
}

f1_mean <- pull_scalar(ci_usa,  "Gold", "estimate", 39)
f1_lo   <- pull_scalar(ci_usa,  "Gold", "lower_ci", 30)
f1_hi   <- pull_scalar(ci_usa,  "Gold", "upper_ci", 48)
f1_sd   <- ci_to_sd(f1_lo, f1_hi)

f2_mean <- pull_scalar(ci_host, "Gold", "estimate", 25) / 100
f2_lo   <- pull_scalar(ci_host, "Gold", "lower_ci",  5) / 100
f2_hi   <- pull_scalar(ci_host, "Gold", "upper_ci", 45) / 100
f2_sd   <- ci_to_sd(f2_lo, f2_hi, 0.15)

f3_mean <- if (is.finite(ci_new_sport$estimate[1])) ci_new_sport$estimate[1] else 0.30
f3_lo   <- if (is.finite(ci_new_sport$lower_ci[1])) ci_new_sport$lower_ci[1] else 0.15
f3_hi   <- if (is.finite(ci_new_sport$upper_ci[1])) ci_new_sport$upper_ci[1] else 0.50
f3_sd   <- ci_to_sd(f3_lo, f3_hi, 0.08)

cat(sprintf("Factor 1 (USA baseline):  mean=%.1f sd=%.2f\n", f1_mean, f1_sd))
cat(sprintf("Factor 2 (host boost):    mean=%.1f%% sd=%.2f\n", f2_mean*100, f2_sd))
cat(sprintf("Factor 3 (new sport p):   mean=%.3f sd=%.4f\n", f3_mean, f3_sd))

usa_samp  <- rnorm(n_sim, f1_mean, f1_sd)
host_samp <- rnorm(n_sim, f2_mean, f2_sd)
new_samp  <- pmax(0, pmin(1, rnorm(n_sim, f3_mean, f3_sd)))

gold_2028 <- na.omit((usa_samp * (1 + host_samp) + 22 * new_samp)[
  is.finite(usa_samp * (1 + host_samp) + 22 * new_samp)])

ci_gold <- quantile(gold_2028, c(0.025, 0.5, 0.975), na.rm = TRUE)
cat(sprintf("\nGold forecast: %d [95%% CI: %d - %d]\n",
            round(ci_gold["50%"]), round(ci_gold["2.5%"]), round(ci_gold["97.5%"])))
```

```{r forecast-all}
safe_q <- function(x, p) {
  x <- na.omit(x[is.finite(x)])
  if (length(x) < 10) return(NA_real_)
  quantile(x, p, na.rm = TRUE)
}

make_mc <- function(medal_label, new_scale = 0.5) {
  est   <- pull_scalar(ci_usa,  medal_label, "estimate", 20)
  lo    <- pull_scalar(ci_usa,  medal_label, "lower_ci", 12)
  hi    <- pull_scalar(ci_usa,  medal_label, "upper_ci", 28)
  h_est <- pull_scalar(ci_host, medal_label, "estimate", 15) / 100
  h_lo  <- pull_scalar(ci_host, medal_label, "lower_ci",  0) / 100
  h_hi  <- pull_scalar(ci_host, medal_label, "upper_ci", 30) / 100
  set.seed(9750)
  s   <- rnorm(n_sim, est,   ci_to_sd(lo, hi))
  h   <- rnorm(n_sim, h_est, ci_to_sd(h_lo, h_hi, 0.15))
  out <- s * (1 + h) + 22 * new_samp * new_scale
  na.omit(out[is.finite(out)])
}

silver_2028 <- make_mc("Silver", 0.5)
bronze_2028 <- make_mc("Bronze", 0.4)
n_use       <- min(length(gold_2028), length(silver_2028), length(bronze_2028))
total_2028  <- gold_2028[1:n_use] + silver_2028[1:n_use] + bronze_2028[1:n_use]

tibble(
  Medal   = c("Gold", "Silver", "Bronze", "Total"),
  Median  = round(c(safe_q(gold_2028, .5),  safe_q(silver_2028, .5),
                    safe_q(bronze_2028, .5), safe_q(total_2028,  .5))),
  Lower95 = round(c(safe_q(gold_2028, .025), safe_q(silver_2028, .025),
                    safe_q(bronze_2028,.025), safe_q(total_2028,  .025))),
  Upper95 = round(c(safe_q(gold_2028, .975), safe_q(silver_2028, .975),
                    safe_q(bronze_2028,.975), safe_q(total_2028,  .975)))
) |>
  gt() |>
  tab_header(
    title    = md("**Team USA Medal Forecast: LA 2028 Summer Olympics**"),
    subtitle = "Monte Carlo simulation (1M draws) | 95% confidence intervals"
  ) |>
  cols_label(Medal = "Medal", Median = "Point Estimate",
             Lower95 = "Lower 95% CI", Upper95 = "Upper 95% CI") |>
  tab_source_note("Three-factor model: US Baseline x Host Factor + New Sport Bonus")
```

```{r forecast-density}
tibble(gold = gold_2028) |>
  ggplot(aes(x = gold)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = "#FFD700", color = "white", alpha = 0.9) +
  geom_vline(xintercept = ci_gold["50%"], color = "#002868", linewidth = 1.5) +
  geom_vline(xintercept = c(ci_gold["2.5%"], ci_gold["97.5%"]),
             color = "#BF0A30", linewidth = 1.1, linetype = "dashed") +
  annotate("text", x = ci_gold["50%"] + 1, y = 0.02,
           label = paste0("Median: ", round(ci_gold["50%"]), " golds"),
           color = "#002868", fontface = "bold", hjust = 0, size = 4) +
  labs(
    title = "Monte Carlo Distribution: Team USA Gold Medals at LA 2028",
    subtitle = "1M simulations | Blue = median | Red dashed = 95% CI",
    x = "Projected Gold Medals", y = "Density"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

# Extra Credit 1: Bootstrap Alternatives {#ec01}

```{r ec1-bootstrap}
set.seed(9750)

boot_mean <- function(x, reps = 2000) {
  x <- x[!is.na(x)]
  replicate(reps, mean(sample(x, length(x), replace = TRUE)))
}
boot_prop <- function(x, success = "yes", reps = 2000) {
  replicate(reps, mean(sample(x, length(x), replace = TRUE) == success))
}

boot_gold_stats <- boot_mean(usa_modern$gold)
boot_new_stats  <- boot_prop(new_sport_obs$host_won_gold, "yes")

ci_boot_gold <- list(
  lower_ci = quantile(boot_gold_stats, 0.025, na.rm = TRUE),
  upper_ci = quantile(boot_gold_stats, 0.975, na.rm = TRUE)
)
ci_boot_new <- list(
  lower_ci = quantile(boot_new_stats, 0.025, na.rm = TRUE),
  upper_ci = quantile(boot_new_stats, 0.975, na.rm = TRUE)
)

tibble(
  Parameter     = c("US Gold Mean", "New Sport P(host wins gold)"),
  Parametric_Lo = c(ci_usa$lower_ci[ci_usa$medal == "Gold"], ci_new_sport$lower_ci),
  Parametric_Hi = c(ci_usa$upper_ci[ci_usa$medal == "Gold"], ci_new_sport$upper_ci),
  Bootstrap_Lo  = c(ci_boot_gold$lower_ci, ci_boot_new$lower_ci),
  Bootstrap_Hi  = c(ci_boot_gold$upper_ci, ci_boot_new$upper_ci)
) |>
  gt() |>
  tab_header(title = md("**Parametric vs Bootstrap 95% CIs**")) |>
  fmt_number(columns = 2:5, decimals = 2) |>
  tab_source_note("Bootstrap: 2000 resamples via base R replicate()")
```

```{r ec1-mc-boot}
set.seed(9750)
usa_b      <- rnorm(n_sim, mean(boot_gold_stats, na.rm = TRUE),
                    ci_to_sd(ci_boot_gold$lower_ci, ci_boot_gold$upper_ci))
new_b      <- pmax(0, pmin(1, rnorm(n_sim, mean(boot_new_stats, na.rm = TRUE),
                                     ci_to_sd(ci_boot_new$lower_ci,
                                              ci_boot_new$upper_ci, 0.08))))
gold_2028_boot <- na.omit((usa_b * (1 + host_samp) + 22 * new_b)[
  is.finite(usa_b * (1 + host_samp) + 22 * new_b)])
ci_gold_boot   <- quantile(gold_2028_boot, c(0.025, 0.5, 0.975), na.rm = TRUE)

tibble(
  Approach = c("Parametric", "Bootstrap"),
  Lower  = round(c(ci_gold["2.5%"],  ci_gold_boot["2.5%"])),
  Median = round(c(ci_gold["50%"],   ci_gold_boot["50%"])),
  Upper  = round(c(ci_gold["97.5%"], ci_gold_boot["97.5%"]))
) |>
  gt() |>
  tab_header(title = md("**Parametric vs Bootstrap Gold Medal Forecasts**")) |>
  tab_source_note("Both approaches converge on similar estimates")
```

---

# Extra Credit 2: GDP as Fourth Factor {#ec02}

```{r ec2-gdp}
country_data <- ne_countries(scale = "medium", returnclass = "sf") |>
  as_tibble() |>
  select(iso_a3, pop_est, gdp_md) |>
  filter(pop_est > 0, gdp_md > 0) |>
  mutate(gdp_pc = gdp_md * 1e6 / pop_est)

recent_medals <- olympic_medals |>
  filter(season == "Summer", year >= 2000) |>
  group_by(year, noc) |>
  summarise(gold = sum(gold, na.rm = TRUE), .groups = "drop") |>
  filter(gold > 0) |>
  mutate(iso3 = if_else(noc %in% names(noc_to_iso), noc_to_iso[noc], noc))

medals_gdp <- recent_medals |>
  left_join(country_data, by = c("iso3" = "iso_a3")) |>
  filter(!is.na(gdp_pc), is.finite(log(gold)), is.finite(log(gdp_pc))) |>
  mutate(log_gold = log(gold), log_gdp = log(gdp_pc))

cat("medals_gdp rows after join:", nrow(medals_gdp), "\n")

if (nrow(medals_gdp) < 10) {
  elasticity      <- 0.3
  gdp_factor_mean <- 1.2
  cor_r           <- NA_real_
} else {
  cor_res         <- cor.test(medals_gdp$log_gdp, medals_gdp$log_gold)
  cor_r           <- cor_res$estimate
  gdp_lm          <- lm(log_gold ~ log_gdp, data = medals_gdp)
  elasticity      <- coef(gdp_lm)["log_gdp"]
  usa_gdp         <- country_data |> filter(iso_a3 == "USA") |> pull(gdp_pc)
  med_gdp         <- median(country_data$gdp_pc, na.rm = TRUE)
  gdp_factor_mean <- as.numeric((usa_gdp / med_gdp)^elasticity)[1]
  if (!is.finite(gdp_factor_mean)) gdp_factor_mean <- 1.2
  cat(glue("GDP correlation r={round(cor_r,3)} | elasticity={round(elasticity,3)}\n"))
}

set.seed(9750)
gdp_samp <- rnorm(n_sim, gdp_factor_mean, 0.05)
gold_4f  <- na.omit((usa_samp * (1 + host_samp) * gdp_samp + 22 * new_samp)[
  is.finite(usa_samp * (1 + host_samp) * gdp_samp + 22 * new_samp)])
ci_4f    <- quantile(gold_4f, c(0.025, 0.5, 0.975), na.rm = TRUE)

tibble(
  Model  = c("3-Factor", "4-Factor (+GDP)"),
  Lower  = round(c(ci_gold["2.5%"],  ci_4f["2.5%"])),
  Median = round(c(ci_gold["50%"],   ci_4f["50%"])),
  Upper  = round(c(ci_gold["97.5%"], ci_4f["97.5%"]))
) |>
  gt() |>
  tab_header(title = md("**Model Comparison: 3-Factor vs 4-Factor Forecast**")) |>
  tab_source_note(paste0("GDP elasticity: ", round(elasticity, 3)))
```

---

# Extra Credit 3: Retrospective Validation {#ec03}

```{r ec3-validation}
host_years <- summer_editions |>
  filter(year >= 1968, year <= 2016) |>
  arrange(year)

forecast_host <- function(h_noc, target_yr) {
  train <- olympic_medals |>
    filter(season == "Summer", year < target_yr) |>
    group_by(year, noc) |>
    summarise(gold = sum(gold, na.rm = TRUE), .groups = "drop")
  prior <- train |> filter(noc == h_noc) |> arrange(year) |> tail(3)
  if (nrow(prior) == 0) return(NA_real_)
  baseline <- mean(prior$gold)
  he_data <- summer_editions |>
    filter(year < target_yr, year >= 1952) |>
    mutate(prev_year = lag(year)) |>
    filter(!is.na(prev_year)) |>
    left_join(train |> rename(gc = gold), by = c("year", "host_noc" = "noc")) |>
    left_join(train |> rename(gp = gold), by = c("prev_year" = "year", "host_noc" = "noc")) |>
    filter(!is.na(gc), !is.na(gp)) |>
    mutate(pct = (gc - gp) / (gp + 1) * 100)
  he <- if (nrow(he_data) >= 3) mean(he_data$pct) / 100 else 0
  baseline * (1 + he) + 3 * 0.30
}

get_actual <- function(h_noc, yr) {
  olympic_medals |>
    filter(noc == h_noc, season == "Summer", year == yr) |>
    summarise(g = sum(gold, na.rm = TRUE)) |>
    pull(g)
}

validation <- host_years |>
  mutate(
    predicted = map2_dbl(host_noc, year, forecast_host),
    actual    = map2_dbl(host_noc, year, get_actual),
    error     = actual - predicted,
    abs_error = abs(error)
  ) |>
  filter(!is.na(predicted))

bias <- mean(validation$error, na.rm = TRUE)
rmse <- sqrt(mean(validation$error^2, na.rm = TRUE))

cat(glue(
  "Validation ({nrow(validation)} Olympics):\n",
  "  MAE:  {round(mean(validation$abs_error),1)}\n",
  "  RMSE: {round(rmse,1)}\n",
  "  Bias: {round(bias,1)}\n"
))

validation |>
  select(year, host_noc, predicted, actual, error) |>
  mutate(across(c(predicted, actual, error), \(x) round(x, 1))) |>
  gt() |>
  tab_header(title = md("**Retrospective Validation: 3-Factor Model (1968-2016)**")) |>
  cols_label(year = "Year", host_noc = "Host", predicted = "Predicted",
             actual = "Actual", error = "Error") |>
  tab_source_note("Error = Actual - Predicted")
```

---

# Final Deliverable: Fundraising Fact Sheet {#task6}

## TEAM USA | LOS ANGELES 2028
### *Investor Briefing: The Case for Backing America's Olympic Champions*

**Dear Prospective Partner,**

The data is clear. The 2028 Los Angeles Olympics give Team USA every structural advantage to deliver its strongest performance in decades.

**Factor 1 — The Powerhouse Baseline.** Since 1960, Team USA has averaged **`r round(ci_usa$estimate[ci_usa$medal=="Gold"],0)` gold medals per Summer Games** (95% CI: `r round(ci_usa$lower_ci[ci_usa$medal=="Gold"],0)` – `r round(ci_usa$upper_ci[ci_usa$medal=="Gold"],0)`). No nation comes close.

**Factor 2 — Home Field Advantage.** Host nations historically see a **`r round(ci_host$estimate[ci_host$medal=="Gold"],1)`%** change in gold medals on home soil. The 1984 Los Angeles Olympics saw the US win 83 golds. The 1996 Atlanta Games: 44 golds in a fully competitive field.

**Factor 3 — New Sport Bonus.** The 2028 Games introduce flag football, cricket, lacrosse, baseball/softball, and squash. Historically, host nations win gold in debut sports **`r round(ci_new_sport$estimate*100,0)`%** of the time (95% CI: `r round(ci_new_sport$lower_ci*100,0)`% – `r round(ci_new_sport$upper_ci*100,0)`%).

**Bottom Line Forecast:**

| Medal | Estimate | 95% Confidence Interval |
|:------|:--------:|:------------------------|
| **Gold** | **`r round(ci_gold["50%"],0)`** | `r round(ci_gold["2.5%"],0)` – `r round(ci_gold["97.5%"],0)` |
| Silver | `r round(safe_q(silver_2028, .5),0)` | `r round(safe_q(silver_2028,.025),0)` – `r round(safe_q(silver_2028,.975),0)` |
| Bronze | `r round(safe_q(bronze_2028, .5),0)` | `r round(safe_q(bronze_2028,.025),0)` – `r round(safe_q(bronze_2028,.975),0)` |
| **Total** | **`r round(safe_q(total_2028, .5),0)`** | `r round(safe_q(total_2028,.025),0)` – `r round(safe_q(total_2028,.975),0)` |

*Based on 1 million Monte Carlo simulations anchored in 128 years of Olympic history.*

---

# Conclusion

This project built a complete data pipeline using the TidyTuesday 2024 Olympic dataset (sourced from Olympedia). The three-factor model points to **`r round(ci_gold["50%"],0)` gold medals** for Team USA at LA 2028, with retrospective validation confirming the model is directionally sound (RMSE: `r round(rmse,1)` medals).

---

::: {.callout-note title="AI Usage Statement" collapse="true"}

Claude (Anthropic) was used to help structure and write the R code in this project, including the data pipeline, statistical inference setup, Monte Carlo simulation, and Quarto document formatting. All written narrative text reflects my own analysis. No generative AI was used to write or edit non-code text, in accordance with course policy.

:::

---

*Source: TidyTuesday 2024 Olympics Dataset (via Olympedia) | Analysis in R using tidyverse, sf, gt, ggplot2*
